---
title: "Loading Petrographer Models from Pins"
author: "petrographer team"
format:
  html:
    code-fold: true
    toc: true
editor: visual
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(petrographer)
if (requireNamespace("pins", quietly = TRUE)) {
  library(pins)
}
```

## Pins-First Architecture

`train_model()` now automatically pins models to `.petrographer/` with all metadata:

- Model id and version (timestamp-based run id)
- Training artifacts (`model_best.pth`, `config.yaml`, `metrics.json`)
- Training metadata (backbone, freeze depth, batch size, learning rate, etc.)
- Dataset information

Models are versioned automatically via pins. `from_pretrained()` downloads models from the public hub or loads from local pins, caching downloads automatically.

This notebook walks through the workflow.

## Configure a Pins Board

For publishing, we use a writable board rooted in a local folder (e.g. a Dropbox directory that you share read-only on the web). `board_user()` creates it if needed.

```{r}
board <- board_user()
board
```

To serve pretrained models to users, publish the contents of this folder to a shared location (the official petrographer hub uses a Dropbox folder). Consumers can then call `from_pretrained()` with no additional configuration; the package points at the public hub by default.

## Publish an Existing Model Directory

Assume you have a trained model in `Detectron2_Models/<name>`. The training pipeline now writes `manifest.json` automatically, but `publish_model()` will regenerate it if needed.

```{r}
example_dir <- "Detectron2_Models/inclusions_v2"

if (dir.exists(example_dir)) {
  publish_result <- publish_model(
    model_dir = example_dir,
    model_id  = "petrography/inclusions_v2",
    board     = board,
    metadata  = list(notes = "Demo publish from notebook")
  )
  publish_result$manifest[c("model_id", "version", "metadata")]
} else {
  cli::cli_alert_warning("Example directory not found. Run train_model() first or adjust example_dir.")
}
```

After publishing, `publish_model()` writes the model bundle to the pins board.

## Inspect Manifest Without Downloading

`model_info()` reads the manifest from pin metadata when available and, if needed, downloads the pin once to access the manifest on disk.

```{r}
if (exists("publish_result")) {
  manifest_preview <- model_info("petrography/inclusions_v2", board = board)
  str(manifest_preview[c("model_id", "version", "metadata")], max.level = 1)
}
```

## Load a Model From Pretrained

`from_pretrained()` downloads (or reuses cached) weights from the hosted hub and returns a ready-to-use `PetrographyModel` object.

```{r}
if (exists("publish_result")) {
  model <- from_pretrained("petrography/inclusions_v2", device = "cpu")

  # Access model properties
  model$model_path
  model$config_path
  model$manifest$metadata[c("backbone", "freeze_at", "ims_per_batch")]
}
```

The object keeps a reference to the manifest, so you can log or audit the run configuration later.

## Reusing the Local Cache

Subsequent calls to `from_pretrained()` reuse the cached files that `pins::pin_download()` stores on disk. To force a refresh, call `pins::pin_delete()` for the model pin and reload.

## Next Steps

- Models are automatically pinned after each training run to `.petrographer/`
- Use `list_trained_models()` or `list_models(board = "local")` to see your local models
- Load any model with `from_pretrained("model_id")` - it checks local first, then hub
- Force local-only with `from_pretrained("model_id", board = "local")`
- Maintainers can publish models to the hub using `pin_model()` and `pins::write_board_manifest()`

This completes the tour of the pins-first workflow.
