---
title: "Loading Petrographer Models from Pretrained Bundles"
author: "petrographer team"
format:
  html:
    code-fold: true
    toc: true
editor: visual
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(petrographer)
if (requireNamespace("pins", quietly = TRUE)) {
  library(pins)
}
```

## Why a Manifest + Bundle workflow?

`train_model()` now writes a `manifest.json` alongside the Detectron2 files. The manifest captures:

- Model id and version (timestamp-based run id)
- Paths to artifacts (`model_final.pth`, `model_best.pth`, `config.yaml`, `metrics.json`)
- Training metadata (backbone, freeze depth, batch size, etc.)
- Any custom notes you pass via `model_description`

`pg_model_publish()` bundles those files into a tarball, uploads it to your writable board (typically a Dropbox folder), and stores the manifest as pin metadata. `pg_install_pretrained()` mirrors the HuggingFace-style `from_pretrained()` flow: it downloads the bundle on first use, unpacks it into `~/.cache/petrographer/models/<model-id>/<pin-version>/`, reuses cached copies afterwards, and returns paths + metadata.

This notebook walks through that workflow.

## Configure a Pins Board

For publishing, we use a writable board rooted in a local folder (e.g. a Dropbox directory that you share read-only on the web). `pg_board_user()` creates it if needed.

```{r}
board <- pg_board_user()
board
```

To serve pretrained models to users, publish the contents of this folder to a URL (for example, a Dropbox “view-only” link) and set `PETRO_PINS_URL` to that URL. Consumers then call `pg_install_pretrained()` with no additional configuration.

## Publish an Existing Model Directory

Assume you have a trained model in `Detectron2_Models/<name>`. The training pipeline now writes `manifest.json` automatically, but `pg_model_publish()` will regenerate it if needed.

```{r}
example_dir <- "Detectron2_Models/inclusions_v2"

if (dir.exists(example_dir)) {
  publish_result <- pg_model_publish(
    model_dir = example_dir,
    model_id  = "petrography/inclusions_v2",
    board     = board,
    metadata  = list(notes = "Demo publish from notebook")
  )
  publish_result$manifest[c("model_id", "version", "metadata")]
} else {
  cli::cli_alert_warning("Example directory not found. Run train_model() first or adjust example_dir.")
}
```

After publishing, `pg_model_publish()` automatically refreshes the board manifest so a hosted `board_url()` can discover new versions.

## Inspect Manifest Without Downloading

`pg_model_info()` reads the manifest from pin metadata when available and falls back to downloading the bundle if necessary.

```{r}
if (exists("publish_result")) {
  manifest_preview <- pg_model_info("petrography/inclusions_v2", board = board)
  str(manifest_preview[c("model_id", "version", "metadata")], max.level = 1)
}
```

## Load a Model From Pretrained

With `PETRO_PINS_URL` set, `pg_install_pretrained()` downloads (or reuses cached) weights and returns resolved paths plus metadata. `load_model(model_name = …)` delegates to it automatically.

```{r}
if (exists("publish_result")) {
  Sys.setenv(PETRO_PINS_URL = Sys.getenv("PETRO_PINS_URL", ""))
  resolved <- pg_install_pretrained("petrography/inclusions_v2")
  resolved$cache_dir
  resolved$manifest$metadata$training_mode
}
```

Now instantiate a SAHI-backed `PetrographyModel` directly from the pin:

```{r}
if (exists("publish_result")) {
  model <- load_model(model_name = "petrography/inclusions_v2", device = "cpu")
  model$manifest$metadata[c("backbone", "freeze_at", "ims_per_batch")]
}
```

The object keeps a reference to the manifest and pin metadata, so you can log or audit the run configuration later.

## Reusing the Local Cache

Subsequent calls reuse cached weights. Inspect the cache root with `pg_model_cache_root()`.

```{r}
pg_model_cache_root()
```

```{r}
if (exists("publish_result")) {
  fs::dir_tree(pg_model_cache_root(), recurse = 2)
}
```

To force a refresh use `pins::pin_delete()` followed by another `pg_model_from_pretrained()` call, or manually remove the cache directory.

## Next Steps

- After each training run, call `train_model(..., publish_after_train = TRUE, model_board = pg_board_user())` to save and version models automatically.

## Working with Datasets

Datasets can be published and consumed with the same helpers.

```{r}
pg_dataset_publish(
  dataset_dir = here::here("data/processed/inclusions_sliced"),
  dataset_id = "datasets/inclusions_sliced"
)
```

On another machine (or after clearing cache), install from the hosted URL:

```{r}
Sys.setenv(PETRO_PINS_URL = Sys.getenv("PETRO_PINS_URL", ""))
if (nzchar(Sys.getenv("PETRO_PINS_URL"))) {
  ds <- pg_dataset_from_pretrained("datasets/inclusions_sliced")
  ds$dataset_dir
}
```

- Run `pg_publish_pkgdown_manifest()` (or `pg_model_publish(write_manifest = TRUE)`) to refresh the hosted manifest before redeploying your pkgdown site or Dropbox link.
- Adopt a naming convention (`petrography/<dataset>/<variant>`) so teams can find models quickly via `pins::pin_search()`.

This completes the tour of the manifest-driven pins workflow.
