---
title: "Petrographer Model Zoo with pins"
author: "petrographer team"
format:
  html:
    code-fold: true
toc: true
editor: visual
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(petrographer)
library(pins)
```

## Overview

This notebook shows how to create and maintain a simple "model zoo" using pins:

-   Configure a writable board (e.g., a Dropbox folder)
-   Publish trained petrographer models and metadata
-   Discover available models and versions
-   Load models from a hosted read-only URL
-   Refresh the manifest for web hosting

pins keeps artifacts versioned and caches downloads automatically.

## Configure a Board

Choose a board based on how you share models:

```{r}
# Writable board (defaults to here::here(petrographer-pins))
board_local <- board_user()
board_local
```

```{r}
# Hosted read-only board (public petrographer hub)
# Users access this automatically via from_pretrained()
# For manual access, use pins::board_url() with the hub URL
hub_url <- 'https://www.dropbox.com/scl/fo/u72bs8odigu37xt0znwyu/ANJ7Ge2VpZl--_qp_2DtVkg?rlkey=353og7qhaj2gxh8xzs7gxph88&dl=0/'
board_hub <- board_url(hub_url)
```

## Pin a Trained Model

Pin a model directory that contains `model_final.pth` and `config.yaml` (created by `train_model()`).

```{r}
# Example: pin an existing trained model directory
example_model_dir <- here::here("Detectron2_Models/shell_inclusions_v21/output")  # adjust to your model

publish_model(
    model_dir = example_model_dir,
    model_id = "shell_inclusions_v21",
    board = board_local,
    metadata = list(owner = Sys.info()[["user"]])
)
```

Pin automatically right after training by passing `publish_after_train = TRUE` to `train_model()`:

```{r}
# if (FALSE) {
#   train_model(
#     data_dir = "data/processed/shell_mixed",
#     output_name = "shell_detector_v4",
#     num_classes = 5,
#     gpus = 1,
#     publish_after_train = TRUE,
#     model_board = board_local
#   )
# }
```

## Discover Models

List pins on the board and inspect metadata:

```{r}
available <- pin_list(board_local)
available
```

```{r}
if (length(avail <- available)) {
  info <- pins::pin_meta(board_local, avail[[1]])
  str(info$user)
}
```

Search by text:

```{r}
if (requireNamespace("pins", quietly = TRUE)) {
  res <- pins::pin_search(board_local, "shell")
  if (!is.null(res) && "name" %in% names(res)) unique(res$name) else character()
}
```

## Versioning

Most boards support versions. With `pins`, each write creates a new version (when versioning is enabled on the board).

```{r}
if (requireNamespace("pins", quietly = TRUE) && length(available)) {
  pins::pin_versions(board_local, available[[1]])
}
```

## Load Models from the Hosted Hub

Load models directly from the hosted hub with `from_pretrained()`. This downloads and caches the model automatically.

```{r}
# Load from hub (automatically caches)
mdl <- from_pretrained("shell_detector_v3", device = "cpu")
```

## Use the Model

```{r}
# Example prediction (adjust the path to an existing image)
# if (exists("mdl") && file.exists("data/raw/sample.jpg")) {
#   res <- predict_image("data/raw/sample.jpg", mdl, use_slicing = TRUE)
#   head(res)
# }
```

## Housekeeping

Delete a pin (careful: deletes all versions on boards that don’t isolate versions):

```{r}
if (FALSE && requireNamespace("pins", quietly = TRUE) && length(available) > 0) {
  pins::pin_delete(board_local, available[[1]])
}
```

## Tips

-   Prefer versioned boards (`versioned = TRUE`) so you can reproduce past runs.
-   After publishing new models, sync your board folder (e.g., Dropbox) so the `_pins.yaml` file is updated for board_url() consumers.
-   Store concise training metadata with `publish_model()` to make discovery easier.
