---
title: "Dataset Slicing Example"
format: html
---

Example of slicing a dataset for improved dense detection performance.

## Setup

```{r setup, message=FALSE}
library(tidyverse)
library(fs)
library(here)
library(petrographer)

data_dir <- here("data/processed/inclusions")
```

## Original Dataset

```{r original-diagnostics}
# Check original dataset
validate_dataset(data_dir)

# Diagnose annotations
train_diag <- diagnose_annotations(
  annotation_json = path(data_dir, "train/_annotations.coco.json"),
  image_dir = path(data_dir, "train")
)

val_diag <- diagnose_annotations(
  annotation_json = path(data_dir, "valid/_annotations.coco.json"),
  image_dir = path(data_dir, "valid")
)
```

```{r original-plots}
# Plot annotation distribution
bind_rows(
  tibble(count = train_diag$annos_per_image, split = "train"),
  tibble(count = val_diag$annos_per_image, split = "valid")
) %>%
  ggplot(aes(x = count, fill = split)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = 100, linetype = "dashed", color = "red") +
  annotate("text", x = 100, y = Inf, label = "Default maxDets=100",
           vjust = 1.5, hjust = -0.1, color = "red", size = 3) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Original Dataset - Annotations per Image",
    subtitle = "Many images have >100 annotations",
    x = "Number of annotations",
    y = "Number of images"
  ) +
  theme_minimal()
```

## Slice Dataset

```{r slice}
# Slice the dataset
sliced_dir <- slice_dataset(
  input_dir = data_dir,
  output_dir = here("data/processed/inclusions_sliced"),
  slice_size = 1024,
  overlap = 0.2,
  min_area_ratio = 0.1
)
```

## Sliced Dataset Diagnostics

```{r sliced-diagnostics}
# Diagnose sliced dataset
train_sliced_diag <- diagnose_annotations(
  annotation_json = path(sliced_dir, "train/_annotations.coco.json"),
  image_dir = path(sliced_dir, "train")
)

val_sliced_diag <- diagnose_annotations(
  annotation_json = path(sliced_dir, "valid/_annotations.coco.json"),
  image_dir = path(sliced_dir, "valid")
)
```

```{r sliced-comparison}
# Compare original vs sliced
comparison <- tibble(
  dataset = c("Original", "Sliced"),
  train_images = c(train_diag$n_images, train_sliced_diag$n_images),
  train_annotations = c(train_diag$n_annotations, train_sliced_diag$n_annotations),
  val_images = c(val_diag$n_images, val_sliced_diag$n_images),
  val_annotations = c(val_diag$n_annotations, val_sliced_diag$n_annotations)
) %>%
  mutate(
    train_per_image = train_annotations / train_images,
    val_per_image = val_annotations / val_images
  )

print(comparison)

cat("\n== Summary ==\n")
cat(sprintf("Image increase: %.1fx\n", comparison$train_images[2] / comparison$train_images[1]))
cat(sprintf("Annotations per image: %.1f → %.1f\n",
            comparison$train_per_image[1], comparison$train_per_image[2]))
```

```{r sliced-plots}
# Plot comparison
bind_rows(
  tibble(count = train_diag$annos_per_image, dataset = "Original"),
  tibble(count = train_sliced_diag$annos_per_image, dataset = "Sliced")
) %>%
  ggplot(aes(x = count, fill = dataset)) +
  geom_histogram(bins = 30, alpha = 0.6, position = "identity") +
  geom_vline(xintercept = 100, linetype = "dashed", color = "red") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Original vs Sliced - Annotations per Image",
    subtitle = "Slicing reduces annotations per image and increases total images",
    x = "Number of annotations per image",
    y = "Number of images"
  ) +
  theme_minimal()

# Object size comparison
bind_rows(
  tibble(area = train_diag$bbox_areas, dataset = "Original"),
  tibble(area = train_sliced_diag$bbox_areas, dataset = "Sliced")
) %>%
  ggplot(aes(x = log10(area), fill = dataset)) +
  geom_histogram(bins = 50, alpha = 0.6, position = "identity") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Object Size Distribution",
    x = "log10(Area) in pixels²",
    y = "Count"
  ) +
  theme_minimal()
```

## Train on Sliced Dataset

```{r train-sliced, eval=FALSE}
# Train model on sliced data
model_path <- train_model(
  data_dir = sliced_dir,
  output_name = "inclusions_sliced",
  ims_per_batch = 12,
  num_workers = 12,
  hpc_cpus_per_task = 16,
  hpc_mem = "96gb",
  eval_period = 200,
  max_iter = 6000,
  num_classes = 1
)
```

## Summary

**Expected improvements from slicing:**
- More training samples (~2x images)
- Better small object detection
- Lower annotations per image (easier for model)
- Better gradient flow

**When to use slicing:**
- Images >1024px
- Dense detection (100+ objects per image)
- Small objects in large images

**When NOT to slice:**
- All images already <1024px
- Few objects per image (<10)
- Objects are very large
