---
title: "Load and Publish Models with Pins"
author: "petrographer team"
format:
  html:
    toc: true
    code-fold: true
editor: visual
---

This notebook covers two related workflows:

- loading an existing model from the local board or public hub
- publishing a completed local model pin to another board as a maintainer

The package now treats `manifest.json` and `training_summary.json` as the
stable model contract. Those files travel with each pin alongside the model
weights and any upstream RF-DETR artifacts.

```{r setup, message=FALSE, warning=FALSE}
library(petrographer)
library(fs)
library(here)
library(jsonlite)

if (requireNamespace("pins", quietly = TRUE)) {
  library(pins)
}
```

## Local and Hub Loading

`from_pretrained()` checks the local model board first and then falls back to
the public hub when `board = NULL`.

```{r list-local}
list_trained_models()
```

```{r load-local, eval=FALSE}
model <- from_pretrained(
  model_id = "inclusions_small",
  board = "local",
  device = "cpu",
  confidence = 0.5
)

model
```

If you omit `board = "local"`, the package uses its normal smart resolution:
local first, then hub.

```{r load-smart, eval=FALSE}
model <- from_pretrained("inclusions_small", device = "cpu")
```

## Inspect the Loaded Contract

The returned `PetrographyModel` keeps the parsed manifest and training summary
on the R side, so you can inspect provenance without digging through raw files.

```{r inspect-model, eval=FALSE}
names(model$manifest)
model$manifest$model
model$manifest$categories
model$training_summary$training
```

For quick pin metadata, `model_info()` reads the pin metadata recorded on the
board itself.

```{r model-info, eval=FALSE}
model_info("inclusions_small", board = "local")
```

## Look at the Files Behind a Pin

If you want to inspect the downloaded files directly, use `pin_download()`.

```{r inspect-files, eval=FALSE}
local_board <- pins::board_folder(here(".petrographer/models"), versioned = TRUE)
files <- pins::pin_download(local_board, "inclusions_small")
basename(files)
```

```{r inspect-json, eval=FALSE}
manifest_path <- files[grepl("manifest\\.json$", files)][1]
summary_path <- files[grepl("training_summary\\.json$", files)][1]

manifest <- jsonlite::read_json(manifest_path)
training_summary <- jsonlite::read_json(summary_path)

manifest$model
training_summary$final_metrics
```

## Publish a Local Model to Another Board

`train_model()` already pins successful runs to the local board. Maintainers
only need `pin_model()` when copying a finished local model to a shared board
such as the public hub.

Start by finding the local pin contents:

```{r locate-local-pin, eval=FALSE}
local_board <- pins::board_folder(here(".petrographer/models"), versioned = TRUE)
files <- pins::pin_download(local_board, "inclusions_small")
model_dir <- fs::path_dir(files[[1]])
model_dir
```

Then define the destination board. For a folder board:

```{r destination-board, eval=FALSE}
shared_board <- pins::board_folder("~/shared/petrographer-models", versioned = TRUE)
```

Or for a hosted/static board URL:

```{r hub-board, eval=FALSE}
hub_board <- pins::board_url("https://example.com/petrographer/models/")
```

Publish the model directory:

```{r publish, eval=FALSE}
pin_model(
  model_dir = model_dir,
  model_id = "inclusions_small",
  board = shared_board,
  metadata = list(
    notes = "Published from model_from_pretrained.qmd"
  )
)
```

For `board_url()` consumers, maintainers should also refresh the board
manifest after publishing:

```{r board-manifest, eval=FALSE}
pins::write_board_manifest(shared_board)
```

## Refreshing and Reusing Cached Files

Repeated calls to `from_pretrained()` reuse the files downloaded by
`pins::pin_download()`. To inspect versions or clear a cached pin from a local
folder board:

```{r versions, eval=FALSE}
pins::pin_versions(local_board, "inclusions_small")
```

```{r delete-local, eval=FALSE}
# Use carefully: deletes the local pin from this board.
pins::pin_delete(local_board, "inclusions_small")
```

## Next Steps

- Use `list_trained_models()` to browse local runs
- Use `from_pretrained()` to load a model from local storage or the public hub
- Inspect `model$manifest` and `model$training_summary` for provenance
- Maintainers can republish local pins with `pin_model()` and refresh the
  destination board manifest with `pins::write_board_manifest()`
