---
title: "Train RF-DETR Model"
format:
  html:
    toc: true
params:
  task: "detection"           # detection | segmentation
  dataset_id: "inclusions"
  data_dir: "data/processed/inclusions_sliced"
  dataset_description: "Pinned training dataset"
  pin_dataset_first: true
  model_id: null
  model_variant: "small"      # detection: nano/small/medium/large; segmentation: seg_nano/seg_small/seg_medium/seg_large/seg_xlarge/seg_2xlarge/seg_preview
  epochs: 100
  batch_size: "auto"
  time_hours: 4
  validate_every: 2
  early_stopping_patience: null
  confidence: 0.5
  device: "cpu"
  resolution: null
  slice_size: 640
  overlap: 0.2
  max_dets: 300
  test_image_index: 1
  use_hpc: false
  hpc_host: null
  hpc_base_dir: null
---

Train one RF-DETR model at a time with a shared workflow for detection and
segmentation. Set `params$task` and `params$model_variant` to choose the path:

- `task = "detection"` for boxes plus SAHI sliced inference and COCO evaluation
- `task = "segmentation"` for masks plus morphology analysis

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(fs)
library(here)
library(petrographer)

task <- match.arg(params$task, c("detection", "segmentation"))

valid_variants <- if (identical(task, "detection")) {
  c("nano", "small", "medium", "large")
} else {
  c("seg_nano", "seg_small", "seg_medium", "seg_large",
    "seg_xlarge", "seg_2xlarge", "seg_preview")
}

if (!params$model_variant %in% valid_variants) {
  stop(
    sprintf(
      "model_variant '%s' is not valid for task '%s'. Valid choices: %s",
      params$model_variant,
      task,
      paste(valid_variants, collapse = ", ")
    )
  )
}

if (isTRUE(params$use_hpc)) {
  if (!requireNamespace("hipergator", quietly = TRUE)) {
    stop("The hipergator package is required when use_hpc = TRUE.")
  }
  hipergator::hpg_configure(
    host = params$hpc_host,
    base_dir = params$hpc_base_dir
  )
}

data_dir <- here(params$data_dir)
model_id <- params$model_id %||% paste(params$dataset_id, params$model_variant, sep = "_")
out_dir <- here("results", model_id)
dir_create(out_dir)
```

## Dataset

Validate the dataset. If `pin_dataset_first = TRUE`, the notebook also pins the
dataset so training stays reproducible and the model manifest can record the
dataset id/version used for the run.

```{r dataset}
validate_dataset(data_dir)

if (isTRUE(params$pin_dataset_first)) {
  pin_dataset(
    data_dir = data_dir,
    dataset_id = params$dataset_id,
    metadata = list(
      description = params$dataset_description
    )
  )
}
```

```{r preview-annotations, fig.cap="Training sample with ground-truth annotations"}
train_images <- dir_ls(path(data_dir, "train"), glob = "*.jpg")
if (!length(train_images)) {
  train_images <- dir_ls(path(data_dir, "train"), glob = "*.png")
}

stopifnot(length(train_images) > 0)

pg_plot_annotations(
  image_path = train_images[[1]],
  annotation_json = path(data_dir, "train", "_annotations.coco.json")
)
```

## Training

`train_model()` auto-dispatches between local and HPC training. The trained
model is pinned automatically to `.petrographer/models/` and now writes a
stable `manifest.json` and `training_summary.json` alongside the RF-DETR
artifacts.

```{r train}
model_id <- train_model(
  dataset_id = params$dataset_id,
  model_id = model_id,
  model_variant = params$model_variant,
  resolution = params$resolution,
  epochs = params$epochs,
  batch_size = params$batch_size,
  time_hours = params$time_hours,
  validate_every = params$validate_every,
  early_stopping_patience = params$early_stopping_patience
)
```

## Training Curves

`evaluate_training()` prefers the petrographer-owned `training_summary.json`
and falls back to upstream RF-DETR logs only if needed.

```{r curves}
eval_result <- evaluate_training(model_id)

have_train <- nrow(eval_result$training_data) > 0 &&
  "epoch" %in% names(eval_result$training_data)

if (have_train) {
  main_losses <- c("loss", "loss_ce", "loss_bbox", "loss_giou")

  train_long <- eval_result$training_data |>
    select(epoch, any_of(main_losses)) |>
    pivot_longer(-epoch, names_to = "loss_type", values_to = "value") |>
    mutate(split = "train")

  val_long <- if (!is.null(eval_result$validation_data)) {
    eval_result$validation_data |>
      select(epoch, any_of(main_losses)) |>
      pivot_longer(-epoch, names_to = "loss_type", values_to = "value") |>
      mutate(split = "validation")
  } else tibble::tibble()

  bind_rows(train_long, val_long) |>
    filter(!is.na(value)) |>
    ggplot(aes(epoch, value, color = split)) +
    geom_line() +
    facet_wrap(~loss_type, scales = "free_y") +
    labs(title = paste("Training curves -", model_id), y = "loss") +
    theme_minimal()
}
```

```{r val-ap, eval=!is.null(eval_result$validation_data)}
ap_cols <- c("ap", "ap50", "ap75", "map",
             "mAP_50_95", "mAP_50", "F1", "ema_mAP_50_95")

eval_result$validation_data |>
  select(epoch, any_of(ap_cols)) |>
  pivot_longer(-epoch, names_to = "metric", values_to = "value") |>
  filter(!is.na(value)) |>
  ggplot(aes(epoch, value, color = metric)) +
  geom_line() +
  labs(title = "Validation AP / mAP", y = "value") +
  theme_minimal()
```

## Sanity-Check Prediction

```{r predict}
model <- from_pretrained(
  model_id,
  board = "local",
  device = params$device,
  confidence = params$confidence
)

val_images <- dir_ls(path(data_dir, "valid"), glob = "*.jpg")
if (!length(val_images)) {
  val_images <- dir_ls(path(data_dir, "valid"), glob = "*.png")
}

stopifnot(length(val_images) > 0)

test_image <- val_images[[params$test_image_index]]

test_result <- predict_image(
  image_path = test_image,
  model = model,
  use_slicing = identical(task, "detection"),
  slice_size = if (identical(task, "detection")) params$slice_size else NULL,
  overlap = params$overlap,
  output_dir = out_dir
)

cat(sprintf("Test image:       %s\n", path_file(test_image)))
cat(sprintf("Objects detected: %d\n", nrow(test_result)))
if (nrow(test_result) > 0) {
  cat(sprintf("Mean confidence:  %.3f\n", mean(test_result$confidence)))
}
```

```{r gt-overlay, fig.cap="Ground truth"}
pg_plot_annotations(
  image_path = test_image,
  annotation_json = path(data_dir, "valid", "_annotations.coco.json")
)
```

```{r pred-overlay, fig.cap="Prediction"}
pred_image <- test_image |>
  path_file() |>
  path_ext_remove() |>
  paste0("_prediction.png")

knitr::include_graphics(path(out_dir, pred_image))
```

## Detection Evaluation

This section applies only to detection models.

```{r sahi-eval, eval=identical(task, "detection")}
sahi_eval <- evaluate_model_sahi(
  model = model,
  annotation_json = path(data_dir, "valid/_annotations.coco.json"),
  image_dir = path(data_dir, "valid"),
  use_slicing = TRUE,
  slice_size = params$slice_size,
  overlap = params$overlap,
  max_dets = params$max_dets
)

print(sahi_eval$summary, n = Inf)

readr::write_csv(
  sahi_eval$predictions,
  path(out_dir, "sahi_validation_predictions.csv")
)
```

## Segmentation Analysis

This section applies only to segmentation models.

```{r morphology, eval=identical(task, "segmentation") && nrow(test_result) > 0}
test_result |>
  select(class_name, confidence, area, eccentricity, orientation_deg,
         solidity, circularity, size_category, shape_category) |>
  arrange(desc(area)) |>
  head(20) |>
  print()

get_population_stats(test_result)
```

```{r morphology-plots, eval=identical(task, "segmentation") && nrow(test_result) > 0}
test_result |>
  ggplot(aes(log_area, fill = class_name)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  labs(
    title = "Object size distribution by class",
    x = "log10(Area) [pixels^2]",
    y = "Count"
  ) +
  theme_minimal()

test_result |>
  ggplot(aes(eccentricity, circularity, color = class_name, size = area)) +
  geom_point(alpha = 0.6) +
  scale_size_continuous(trans = "log10") +
  labs(
    title = "Shape space (eccentricity x circularity)",
    x = "Eccentricity",
    y = "Circularity"
  ) +
  theme_minimal()
```

```{r segmentation-batch, eval=identical(task, "segmentation")}
seg_batch <- analyze_segmentation_dir(
  input_dir = path(data_dir, "valid"),
  model = model,
  output_dir = path(out_dir, "segmentation_batch")
)

seg_batch$summary
```

## Next Steps

- Compare against prior local runs with `list_trained_models()`
- For detection, review SAHI COCO metrics before publishing
- For segmentation, review morphology summaries across the validation set
- Maintainers can publish curated models to a shared board with `pin_model()`
