---
title: "Model & Dataset Library"
vignette: >
  %\VignetteIndexEntry{Model & Dataset Library}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    code-fold: false
execute:
  eval: true
---

This page is the website-facing catalog for the petrographer model and dataset
library. It is designed for browsing what is available, checking high-level
training details, and copying the right `from_pretrained()` or `train_model()`
calls into your own workflow.

```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
library(petrographer)
library(reactable)
library(htmltools)
library(dplyr)
library(tidyr)
library(purrr)
library(pins)
library(here)

# Connect to S3 hub boards
model_board <- board_url('https://flmnh-ai.s3.us-east-1.amazonaws.com/.petrographer/models/')
dataset_board <- board_url('https://flmnh-ai.s3.us-east-1.amazonaws.com/.petrographer/datasets/')

# Get pins from both boards, but degrade gracefully if the hub is unavailable.
model_pins <- tryCatch(pin_list(model_board), error = function(e) character(0))
dataset_pins <- tryCatch(pin_list(dataset_board), error = function(e) character(0))
hub_available <- length(model_pins) > 0 || length(dataset_pins) > 0

# Create a list of pins with their board type (don't store board object in tibble)
pins_with_board <- bind_rows(
  tibble(name = model_pins, board_type = "model"),
  tibble(name = dataset_pins, board_type = "dataset")
)
```

## Available Models

```{r models}
#| echo: false
if (hub_available && nrow(pins_with_board) > 0) {
  # Get metadata for all pins
  pins_data <- pmap_dfr(pins_with_board, function(name, board_type) {
    # Resolve board based on type
    board <- if (board_type == "model") {
      model_board
    } else {
      dataset_board
    }

    tryCatch({
      meta <- pins::pin_meta(board, name)
      manifest <- meta$user
      catalog <- manifest$catalog_summary %||% manifest
      final_metrics <- catalog$final_metrics %||% manifest$metrics$validation %||% list()

      tibble(
        id = name,
        type = board_type,
        created = catalog$created %||% manifest$created %||% NA_character_,

        # Model-specific fields (RF-DETR)
        model_variant = catalog$model_variant %||% manifest$model_variant %||% NA_character_,
        num_classes = catalog$num_classes %||% manifest$num_classes %||% NA_integer_,
        dataset_id = catalog$dataset_id %||% manifest$dataset_id %||% NA_character_,
        epochs = catalog$epochs %||% manifest$epochs %||% NA_integer_,
        batch_size = catalog$batch_size %||% manifest$batch_size %||% NA_integer_,
        grad_accum_steps = catalog$grad_accum_steps %||% manifest$grad_accum_steps %||% NA_integer_,
        learning_rate = catalog$learning_rate %||% manifest$learning_rate %||% NA_real_,
        training_mins = catalog$training_duration_mins %||% manifest$training_duration_mins %||% NA_real_,
        training_mode = catalog$training_mode %||% manifest$training_mode %||% NA_character_,

        # Performance metrics for catalog display
        map_50_95 = final_metrics$ap_50_95 %||% final_metrics$mAP_50_95 %||% final_metrics$map %||% final_metrics$ap %||% NA_real_,
        map_50 = final_metrics$ap_50 %||% final_metrics$mAP_50 %||% final_metrics$ap50 %||% NA_real_,

        # Dataset-specific fields
        description = manifest$description %||% "",
        num_images = manifest$num_images %||% NA_integer_,

        # Store full manifest for details
        full_manifest = list(manifest)
      )
    }, error = function(e) {
      # Return tibble with all columns to match structure
      tibble(
        id = name,
        type = board_type,
        created = NA_character_,
        model_variant = NA_character_,
        num_classes = NA_integer_,
        dataset_id = NA_character_,
        epochs = NA_integer_,
        batch_size = NA_integer_,
        grad_accum_steps = NA_integer_,
        learning_rate = NA_real_,
        training_mins = NA_real_,
        training_mode = NA_character_,
        map_50_95 = NA_real_,
        map_50 = NA_real_,
        description = "",
        num_images = NA_integer_,
        full_manifest = list(NULL)
      )
    })
  })

  # Filter models
  models <- pins_data %>%
    filter(type == "model" | is.na(type))

  if (nrow(models) > 0) {
    reactable(
      models,
      searchable = TRUE,
      filterable = TRUE,
      defaultPageSize = 10,
      defaultSorted = "map_50_95",
      defaultSortOrder = "desc",
      columns = list(
        id = colDef(
          name = "Model ID",
          minWidth = 180,
          cell = function(value) {
            htmltools::code(
              style = "background: #f5f5f5; padding: 4px 8px; border-radius: 4px; font-size: 13px;",
              value
            )
          }
        ),
        type = colDef(show = FALSE),
        full_manifest = colDef(show = FALSE),
        created = colDef(show = FALSE),
        description = colDef(show = FALSE),
        dataset_id = colDef(show = FALSE),
        epochs = colDef(show = FALSE),
        batch_size = colDef(show = FALSE),
        grad_accum_steps = colDef(show = FALSE),
        learning_rate = colDef(show = FALSE),
        training_mode = colDef(show = FALSE),
        map_50 = colDef(show = FALSE),
        num_images = colDef(show = FALSE),

        model_variant = colDef(
          name = "Variant",
          minWidth = 100,
          cell = function(value) {
            if (is.na(value)) return("")
            htmltools::span(style = "font-size: 13px; color: #555;", value)
          }
        ),
        num_classes = colDef(
          name = "Classes",
          minWidth = 70,
          align = "center",
          cell = function(value) {
            if (is.na(value)) return("")
            htmltools::span(style = "font-weight: 500; color: #666;", value)
          }
        ),
        map_50_95 = colDef(
          name = "mAP 50:95",
          minWidth = 100,
          align = "right",
          cell = function(value) {
            if (is.na(value)) return("—")
            pct <- value * 100
            color <- if (pct >= 50) "#28a745" else if (pct >= 30) "#ffc107" else "#dc3545"
            htmltools::span(
              style = sprintf("font-weight: 700; font-size: 15px; color: %s;", color),
              sprintf("%.1f%%", pct)
            )
          }
        ),
        training_mins = colDef(
          name = "Training Time",
          minWidth = 110,
          align = "right",
          cell = function(value) {
            if (is.na(value)) return("")
            mins <- round(value)
            if (mins < 60) {
              htmltools::span(style = "font-size: 13px; color: #666;", sprintf("%d min", mins))
            } else {
              hours <- floor(mins / 60)
              remaining_mins <- mins %% 60
              htmltools::span(style = "font-size: 13px; color: #666;", sprintf("%dh %dm", hours, remaining_mins))
            }
          }
        )
      ),
      details = function(index) {
        model <- models[index, ]
        manifest <- model$full_manifest[[1]]
        # Handle NULL manifest from error handler
        if (is.null(manifest)) {
          return(htmltools::div(
            style = "padding: 20px; background: #f8f9fa;",
            htmltools::div(
              class = "alert alert-warning",
              "No details available for this model."
            )
          ))
        }

        catalog <- manifest$catalog_summary %||% manifest
        validation <- catalog$final_metrics %||% manifest$metrics$validation %||% list()

        htmltools::div(
          style = "padding: 20px; background: #f8f9fa;",

          # Quick load code at top
          htmltools::div(
            style = "margin-bottom: 20px; padding: 12px; background: white; border-radius: 6px; border-left: 4px solid #2c5aa0;",
            htmltools::h5(style = "margin: 0 0 8px 0; color: #333; font-size: 14px;", "📦 Load Model"),
            htmltools::pre(
              style = "margin: 0; background: #f8f9fa; padding: 10px; border-radius: 4px; overflow-x: auto; position: relative;",
              htmltools::code(
                style = "font-size: 13px; color: #333; font-family: 'Monaco', monospace;",
                sprintf('model <- from_pretrained("%s")\nresults <- predict(model, "image.jpg")', model$id)
              ),
              htmltools::tags$button(
                onclick = sprintf("navigator.clipboard.writeText('model <- from_pretrained(\"%s\")')", model$id),
                class = "btn btn-sm btn-outline-primary",
                style = "position: absolute; top: 6px; right: 6px; padding: 2px 8px; font-size: 11px;",
                "Copy"
              )
            )
          ),

          # Validation Metrics
          if (length(validation) > 0) {
            htmltools::div(
              style = "margin-bottom: 16px;",
              htmltools::h5(style = "margin: 0 0 10px 0; color: #333; font-size: 14px; font-weight: 600;", "📊 Validation Metrics"),
              htmltools::div(
                style = "display: grid; grid-template-columns: repeat(auto-fit, minmax(150px, 1fr)); gap: 10px;",
                lapply(names(validation), function(metric) {
                  value <- validation[[metric]]
                  if (is.numeric(value) && !is.na(value) && metric != "epoch") {
                    label <- gsub("/", " ", toupper(metric))
                    display_val <- if (grepl("ap", tolower(metric))) {
                      sprintf("%.1f%%", value * 100)
                    } else {
                      sprintf("%.3f", value)
                    }

                    htmltools::div(
                      style = "background: white; padding: 10px; border-radius: 4px; border: 1px solid #dee2e6;",
                      htmltools::div(style = "font-size: 11px; color: #6c757d; margin-bottom: 4px;", label),
                      htmltools::div(style = "font-size: 16px; font-weight: 600; color: #2c5aa0;", display_val)
                    )
                  }
                })
              )
            )
          },

          # Training Configuration
          htmltools::div(
            style = "margin-bottom: 16px;",
            htmltools::h5(style = "margin: 0 0 10px 0; color: #333; font-size: 14px; font-weight: 600;", "⚙️ Training Configuration"),
            htmltools::div(
            style = "background: white; padding: 12px; border-radius: 4px; display: grid; grid-template-columns: repeat(auto-fit, minmax(200px, 1fr)); gap: 8px 16px; font-size: 13px;",
              htmltools::div(htmltools::strong("Dataset: "), catalog$dataset_id %||% manifest$dataset_id %||% "—"),
              htmltools::div(htmltools::strong("Task: "), catalog$task %||% "—"),
              htmltools::div(htmltools::strong("Model Variant: "), catalog$model_variant %||% manifest$model_variant %||% "—"),
              htmltools::div(htmltools::strong("Epochs: "), format(catalog$epochs %||% manifest$epochs %||% 0, big.mark = ",")),
              htmltools::div(htmltools::strong("Batch Size: "), catalog$batch_size %||% manifest$batch_size %||% "—"),
              htmltools::div(htmltools::strong("Grad Accum Steps: "), catalog$grad_accum_steps %||% manifest$grad_accum_steps %||% "—"),
              htmltools::div(htmltools::strong("Learning Rate: "),
                if (!is.null(catalog$learning_rate) && is.numeric(catalog$learning_rate))
                  sprintf("%.2e", catalog$learning_rate) else if (!is.null(manifest$learning_rate) && is.numeric(manifest$learning_rate))
                    sprintf("%.2e", manifest$learning_rate) else "—"
              ),
              htmltools::div(htmltools::strong("Training Mode: "), catalog$training_mode %||% manifest$training_mode %||% "—"),
              htmltools::div(htmltools::strong("Duration: "),
                if (!is.na(model$training_mins)) {
                  mins <- round(model$training_mins)
                  if (mins < 60) sprintf("%d min", mins) else sprintf("%.1f hr", mins/60)
                } else "—"
              )
            )
          ),

          # Metadata
          htmltools::div(
            htmltools::h5(style = "margin: 0 0 10px 0; color: #333; font-size: 14px; font-weight: 600;", "ℹ️ Metadata"),
            htmltools::div(
            style = "background: white; padding: 12px; border-radius: 4px; font-size: 13px;",
            htmltools::div(htmltools::strong("Created: "),
              if (!is.null(catalog$created) && !is.na(catalog$created)) {
                format(as.POSIXct(catalog$created), "%Y-%m-%d %H:%M")
              } else if (!is.null(manifest$created) && !is.na(manifest$created)) {
                format(as.POSIXct(manifest$created), "%Y-%m-%d %H:%M")
                } else {
                  "—"
                }
              ),
              htmltools::div(htmltools::strong("Run ID: "), catalog$run_id %||% manifest$run_id %||% manifest$version %||% "—")
            )
          )
        )
      }
    )
  } else {
    htmltools::div(
      class = "alert alert-info",
      "No models available in the hub yet."
    )
  }
} else {
  htmltools::div(
    class = "alert alert-warning",
    "Unable to connect to the model hub right now. The page will show available content only when the hub listing endpoints respond."
  )
}
```

## Available Datasets

```{r datasets}
#| echo: false
if (nrow(pins_with_board) > 0 && exists("pins_data")) {
  # Filter datasets
  datasets <- pins_data %>%
    filter(type == "dataset")

  if (nrow(datasets) > 0) {
    reactable(
      datasets,
      searchable = TRUE,
      filterable = TRUE,
      defaultPageSize = 10,
      columns = list(
        id = colDef(
          name = "Dataset ID",
          minWidth = 200,
          cell = function(value) {
            htmltools::code(
              style = "background: #f5f5f5; padding: 4px 8px; border-radius: 4px; font-size: 13px;",
              value
            )
          }
        ),
        type = colDef(show = FALSE),
        full_manifest = colDef(show = FALSE),
        model_variant = colDef(show = FALSE),
        num_classes = colDef(show = FALSE),
        dataset_id = colDef(show = FALSE),
        epochs = colDef(show = FALSE),
        batch_size = colDef(show = FALSE),
        grad_accum_steps = colDef(show = FALSE),
        learning_rate = colDef(show = FALSE),
        training_mins = colDef(show = FALSE),
        training_mode = colDef(show = FALSE),
        map_50_95 = colDef(show = FALSE),
        map_50 = colDef(show = FALSE),

        description = colDef(
          name = "Description",
          minWidth = 250
        ),
        num_images = colDef(
          name = "Images",
          minWidth = 80,
          align = "center",
          cell = function(value) {
            if (is.na(value)) return("—")
            htmltools::span(style = "font-weight: 500; color: #666;", format(value, big.mark = ","))
          }
        ),
        created = colDef(
          name = "Created",
          minWidth = 120,
          cell = function(value) {
            if (is.na(value)) return("")
            date_part <- sub("T.*", "", value)
            htmltools::span(style = "font-size: 13px; color: #666;", date_part)
          }
        )
      ),
      details = function(index) {
        dataset <- datasets[index, ]
        manifest <- dataset$full_manifest[[1]]

        # Handle NULL manifest from error handler
        if (is.null(manifest)) {
          return(htmltools::div(
            style = "padding: 20px; background: #f8f9fa;",
            htmltools::div(
              class = "alert alert-warning",
              "No details available for this dataset."
            )
          ))
        }

        htmltools::div(
          style = "padding: 20px; background: #f8f9fa;",

          # Usage code
          htmltools::div(
            style = "margin-bottom: 20px; padding: 12px; background: white; border-radius: 6px; border-left: 4px solid #28a745;",
            htmltools::h5(style = "margin: 0 0 8px 0; color: #333; font-size: 14px;", "📦 Use Dataset"),
            htmltools::pre(
              style = "margin: 0; background: #f8f9fa; padding: 10px; border-radius: 4px; overflow-x: auto; position: relative;",
              htmltools::code(
                style = "font-size: 13px; color: #333; font-family: 'Monaco', monospace;",
                sprintf('train_model(\n  dataset_id    = "%s",\n  model_id      = "%s_model",\n  model_variant = "small",\n  epochs        = 50,\n  batch_size    = 4\n)',
                  dataset$id, dataset$id
                )
              ),
              htmltools::tags$button(
                onclick = sprintf("navigator.clipboard.writeText('train_model(dataset_id = \"%s\", model_id = \"%s_model\", model_variant = \"small\", epochs = 50, batch_size = 4)')",
                  dataset$id, dataset$id),
                class = "btn btn-sm btn-outline-success",
                style = "position: absolute; top: 6px; right: 6px; padding: 2px 8px; font-size: 11px;",
                "Copy"
              )
            )
          ),

          # Dataset Info
          htmltools::div(
            htmltools::h5(style = "margin: 0 0 10px 0; color: #333; font-size: 14px; font-weight: 600;", "ℹ️ Dataset Information"),
            htmltools::div(
              style = "background: white; padding: 12px; border-radius: 4px; display: grid; grid-template-columns: repeat(auto-fit, minmax(200px, 1fr)); gap: 8px 16px; font-size: 13px;",
              if (!is.null(manifest$description)) htmltools::div(htmltools::strong("Description: "), manifest$description),
              if (!is.null(manifest$num_images)) htmltools::div(htmltools::strong("Images: "), format(manifest$num_images, big.mark = ",")),
              if (!is.null(manifest$num_classes)) htmltools::div(htmltools::strong("Classes: "), manifest$num_classes),
              if (!is.null(manifest$created_by)) htmltools::div(htmltools::strong("Created By: "), manifest$created_by),
              if (!is.null(manifest$created)) htmltools::div(htmltools::strong("Created: "), as.character(manifest$created))
            )
          )
        )
      }
    )
  } else {
    htmltools::div(
      class = "alert alert-info",
      "No datasets available yet."
    )
  }
}
```

## Loading a Model

All models can be loaded with a single line:

```r
library(petrographer)

# Load any model from the hub
model <- from_pretrained("model_id")

# Run predictions
results <- predict(model, "path/to/image.jpg")
```

## Publishing to the Library

Maintainers can publish a finished local model pin to the shared library board
with `pin_model()`. The exact destination board depends on the deployment
environment, so the example below is intentionally generic:

```r
# Download the local pin so you can locate the model directory
local_board <- pins::board_folder(here::here(".petrographer/models"), versioned = TRUE)
files <- pins::pin_download(local_board, "my_model")
model_dir <- fs::path_dir(files[[1]])

# Configure the destination board
# board <- pins::board_url("https://example.com/petrographer/models/")
# board <- pins::board_folder("~/shared/petrographer-models", versioned = TRUE)

# Publish the local pin to that board
pin_model(
  model_dir = model_dir,
  model_id  = "my_model",
  board     = board,
  metadata  = list(
    description   = "Brief description",
    model_variant = "small"
  )
)

# Refresh the destination board manifest for board_url() consumers
pins::write_board_manifest(board)
```

Contact the package maintainers to have your models added to the public hub.
