---
title: "Training Custom Detection Models"
format:
  html:
    toc: true
execute:
  eval: false
  warning: false
  message: false
---

This vignette walks through training custom object detection models for petrographic analysis. It covers both local training (accessible to all users) and HPC training (specific to University of Florida's HiPerGator, but adaptable to other SLURM-based systems).

## Prerequisites

Before training, ensure you have:

1. **Dataset in COCO format** with train/valid splits
2. **Python environment** with detectron2, sahi, torch installed
3. **GPU access** (recommended) - training on CPU is slow but possible

## Dataset Preparation

### Validate Your Dataset

Use `validate_dataset()` to check your dataset structure and COCO format:

```{r validate}
library(petrographer)
library(dplyr)
library(ggplot2)
library(fs)

# Validate dataset structure
validate_dataset("data/processed/my_dataset")
```

Your dataset should have this structure:

```
my_dataset/
├── train/
│   ├── _annotations.coco.json
│   └── *.jpg
└── valid/
    ├── _annotations.coco.json
    └── *.jpg
```

### Optional: Dataset Slicing

For large, heterogeneous image sizes, use SAHI slicing to create uniform tiles:

```{r slice}
# Slice large images into 1024x1024 tiles
slice_dataset(
  input_dir = "data/raw/large_images",
  output_dir = "data/processed/my_dataset_sliced",
  slice_size = 1024,
  overlap = 0.2
)
```

### Optional: Pin Your Dataset

Pin datasets for versioning and easy reference:

```{r pin-dataset}
# Pin dataset to local board
pin_dataset(
  data_dir = "data/processed/my_dataset",
  dataset_id = "my_dataset_v1"
)

# List pinned datasets
list_datasets()
```

## Local Training

Local training runs on your machine (CPU or GPU). This is the simplest way to train a model.

### Basic Training

```{r train-local}
# Train a model locally
model_id <- train_model(
  data_dir = "data/processed/my_dataset",
  model_id = "my_detector_v1",
  num_classes = 3,           # Number of object classes
  max_iter = 12000,          # Training iterations
  device = "cuda"            # "cpu", "cuda", or "mps"
)
```

The model will be automatically pinned to your local board (`.petrographer/`) with full metadata.

### Key Parameters Explained

**Dataset specification:**
```{r dataset-params}
# Option 1: Use a directory path
train_model(data_dir = "data/processed/my_dataset", ...)

# Option 2: Use a pinned dataset ID
train_model(dataset_id = "my_dataset_v1", ...)
```

**Model architecture:**
```{r model-params}
# Backbone options: "resnet50" (default), "resnet101", "resnext101"
backbone = "resnet50"

# Freeze early layers for faster training (0-5)
# Higher = more frozen = faster training, less adaptation
freeze_at = 2  # Freeze stem + res2 (recommended)
```

**Training schedule:**
```{r schedule-params}
# Iterations (depends on dataset size)
max_iter = 12000     # ~1-2 hours on modern GPU

# Learning rate (NULL = auto-scaling based on batch size and freeze_at)
learning_rate = NULL  # Let petrographer compute optimal rate

# Batch size (NA = auto: 2 images per GPU)
ims_per_batch = NA

# Validation frequency
eval_period = 1000    # Evaluate every 1000 iterations
```

### What Happens During Training

1. **Dataset validation** - Checks COCO format and class consistency
2. **Config generation** - Computes batch size, learning rate, workers
3. **Training** - Calls Detectron2 via Python with:
   - WarmupCosineLR schedule
   - Differential learning rates (0.1x backbone, 1.0x head)
   - Data augmentations (flips, rotations, color jitter)
4. **Auto-pinning** - Saves model to `.petrographer/` with metadata

## Loading and Testing Trained Models

### Load a Trained Model

```{r load-trained}
# Load from local board (checks .petrographer/ first, then hub)
model <- from_pretrained("my_detector_v1", device = "cuda")

# Alternatively, force local board
model <- from_pretrained("my_detector_v1", board = "local", device = "cuda")
```

### Test on Validation Images

```{r test-model}
# Get validation images
val_images <- dir_ls("data/processed/my_dataset/valid", glob = "*.jpg")

# Test on a single image
results <- predict(model, val_images[1])

# View detections
results |>
  select(image_id, category_name, confidence, area, perimeter)

# Test on multiple images
batch_results <- predict_images(
  image_dir = "data/processed/my_dataset/valid",
  model = model,
  output_dir = "results/validation_test"
)
```

## Evaluating Training Results

### Parse Training Metrics

```{r eval-training}
# Load metrics from training
eval_result <- evaluate_training(
  model_dir = path(".petrographer", model_id, "current")
)

# Summary statistics
eval_result$summary
```

### Plot Training Curves

```{r plot-training}
# Loss curves
eval_result$training_data |>
  select(iteration, contains("loss")) |>
  tidyr::pivot_longer(-iteration, names_to = "loss_type", values_to = "loss") |>
  filter(!is.na(loss)) |>
  ggplot(aes(iteration, loss, color = loss_type)) +
  geom_line() +
  facet_wrap(~loss_type, scales = "free_y") +
  labs(title = "Training Loss") +
  theme_minimal()

# Validation metrics
eval_result$validation_data |>
  tidyr::pivot_longer(-iteration, names_to = "metric", values_to = "value") |>
  filter(!is.na(value)) |>
  ggplot(aes(iteration, value)) +
  geom_line() +
  geom_point(size = 0.5) +
  facet_wrap(~metric, scales = "free_y") +
  labs(title = "Validation Metrics") +
  theme_minimal()
```

### Evaluate on Validation Set

```{r eval-sahi}
# Run full COCO evaluation with SAHI
sahi_eval <- evaluate_model_sahi(
  model = model,
  annotation_json = "data/processed/my_dataset/valid/_annotations.coco.json",
  image_dir = "data/processed/my_dataset/valid",
  use_slicing = FALSE
)

# COCO metrics
sahi_eval$summary |>
  filter(metric %in% c("AP", "AP50", "AP75", "AR@100"))
```

## HPC Training (Advanced)

::: {.callout-note}
This section is specific to users of **University of Florida's HiPerGator**. However, the code demonstrates patterns that can be adapted to other SLURM-based HPC systems by modifying the `hipergator` package integration.
:::

HPC training is useful for:
- Large datasets requiring long training times
- Multi-GPU training
- Running multiple experiments in parallel

### One-Time Setup

Configure your HPC connection once per session:

```{r hpc-config}
library(hipergator)

# Configure HiPerGator connection
hpg_configure(
  host = "hpg",
  base_dir = "/blue/mygroup/myusername/petrographer"
)
```

You can also set these via `.Renviron`:
```
PETROGRAPHER_HPC_HOST=hpg.rc.ufl.edu
PETROGRAPHER_HPC_BASE_DIR=/blue/mygroup/myusername/petrographer
```

### Run HPC Training

Training mode is auto-detected based on `hipergator` configuration. Use the **same** `train_model()` call:

```{r train-hpc}
# Same call as local training - HPC mode auto-detected!
model_id <- train_model(
  data_dir = "data/processed/my_dataset",
  model_id = "my_detector_hpc_v1",
  num_classes = 3,
  max_iter = 12000
  # No hpc_* params needed - uses hipergator config!
)
```

### Resource Configuration

Customize HPC resources for your job:

```{r hpc-resources}
model_id <- train_model(
  data_dir = "data/processed/my_dataset",
  model_id = "my_large_model",
  num_classes = 5,
  max_iter = 20000,

  # HPC resource hints
  gpus = 2,                    # Multi-GPU training
  hpc_cpus_per_task = 28,      # CPUs (default: 14 per GPU)
  hpc_mem = "48gb",            # Memory (default: 24gb per GPU)

  # Batch size must be divisible by GPU count
  ims_per_batch = 16           # 8 per GPU
)
```

### How HPC Training Works

Behind the scenes, `train_model()`:

1. **Uploads artifacts** to HPC:
   - Dataset → `base_dir/datasets/{dataset_id}/`
   - Training script → `base_dir/scripts/train.py`
   - Both use `rsync` - unchanged files are skipped!

2. **Submits SLURM job** with:
   - Working directory: `base_dir`
   - Command uses relative paths (clean, portable)
   - GPU allocation via `hipergator` resource spec

3. **Monitors job** until completion

4. **Downloads results**:
   - `model_best.pth` (weights)
   - `config.yaml` (model config)
   - `metadata.json` (class names)
   - `metrics.json` (training history)

5. **Pins model** to local board for easy loading

### Shared Directory Structure

HPC uses an efficient shared structure to avoid re-uploading:

```
/blue/base_dir/
├── datasets/
│   └── my_dataset_v1/          # Shared across all models
│       ├── train/
│       └── valid/
├── scripts/
│   └── train.py                # Shared script
└── models/
    └── my_detector_hpc_v1/
        └── 20250113120000/     # Run ID (timestamp)
            └── output/
                ├── model_best.pth
                ├── config.yaml
                └── metadata.json
```

This means:
- First training uploads everything
- Subsequent trainings only upload changed files
- Multiple models can share datasets and scripts

### Monitoring and Troubleshooting

```{r hpc-monitor}
# Check HPC connection
hipergator::hpg_check_connection()

# View your base directory
hipergator::hpg_show_base_dir()

# List running jobs (if you have access)
# hipergator provides job monitoring functions
```

If training fails:
1. Check error messages in the R console
2. Verify `base_dir` exists on HPC
3. Ensure dataset paths are correct
4. Check GPU availability on your partition

## Advanced Topics

### Custom Learning Rates

Override auto-scaling for fine control:

```{r custom-lr}
train_model(
  data_dir = "data/processed/my_dataset",
  model_id = "my_model",
  num_classes = 3,
  learning_rate = 0.0005,  # Manual LR (head rate; backbone gets 0.1x)
  max_iter = 12000
)
```

The default auto-scaling uses:
- Batch size (linear scaling)
- `freeze_at` (more frozen = higher LR)
- Base rate: 0.001 for `freeze_at = 2`

### Backbone Selection

Choose a backbone based on your needs:

```{r backbones}
# ResNet50 (default) - Fast, good for most tasks
backbone = "resnet50"

# ResNet101 - Deeper, better accuracy, slower
backbone = "resnet101"

# ResNeXt101 - Best accuracy, slowest, needs more memory
backbone = "resnext101"
```

### Freeze Strategies

Control which layers to train:

```{r freeze}
# freeze_at = 0: Train entire backbone (slow, best adaptation)
# freeze_at = 1: Freeze stem only (good balance)
# freeze_at = 2: Freeze stem + res2 (default, faster)
# freeze_at = 3: Freeze stem + res2 + res3
# freeze_at = 4: Freeze stem + res2 + res3 + res4
# freeze_at = 5: Freeze entire backbone (fastest, least adaptation)

train_model(
  ...,
  freeze_at = 1,  # Less frozen = slower but better domain adaptation
  learning_rate = NULL  # Auto-scale adjusts for freeze_at
)
```

### Using Pinned Datasets

Pin datasets once, reuse many times:

```{r use-pinned}
# Pin dataset
pin_dataset("data/processed/my_dataset", "my_dataset_v1")

# Train multiple models with same dataset
train_model(dataset_id = "my_dataset_v1", model_id = "model_a", ...)
train_model(dataset_id = "my_dataset_v1", model_id = "model_b", ...)
```

## Publishing to Hub (Maintainers)

To share models with the community, maintainers can publish to the public hub:

```{r publish}
# Create hub board (maintainers only)
hub_board <- pins::board_folder(
  here::here("pkgdown/assets/pins"),
  versioned = TRUE
)

# Pin model to hub
pin_model(
  model_dir = ".petrographer/my_detector_v1/current",
  model_id = "my_detector_v1",
  board = hub_board,
  metadata = list(
    description = "Shell detector optimized for XPL images",
    backbone = "resnet50",
    num_classes = 5
  )
)

# Update manifest
pins::write_board_manifest(hub_board)

# Deploy via pkgdown
# pkgdown::build_site()
# git commit & push
```

See `data-raw/publish-to-hub.R` in the package source for the complete workflow.

## Next Steps

- **Prediction workflow:** See `vignette("whole-slide-basics")` for inference
- **Model library:** See `vignette("model-library")` for available pretrained models
- **Function reference:** See `?train_model`, `?from_pretrained`, `?evaluate_model_sahi`

## Troubleshooting

**Training fails with CUDA out of memory:**
- Reduce `ims_per_batch` (e.g., try 4 or 2)
- Use a smaller `backbone` (resnet50 instead of resnext101)
- Increase `freeze_at` to train fewer parameters

**Training is very slow:**
- Ensure `device = "cuda"` (not "cpu")
- Check GPU availability: `reticulate::py_run_string("import torch; print(torch.cuda.is_available())")`
- Reduce `eval_period` (evaluate less frequently)

**Class names not showing in predictions:**
- Ensure your COCO JSON has a `categories` list with `name` fields
- Check `metadata.json` was created: `list.files(".petrographer/my_model/current")`
- The model should load `category_mapping` automatically

**HPC job fails:**
- Verify `base_dir` exists: `hipergator::hpg_show_base_dir()`
- Check dataset paths are correct relative to `data_dir`
- Ensure GPU partition has availability
