---
title: "Whole-Slide Prediction Basics"
format:
  html:
    toc: true
execute:
  warning: false
  message: false
  eval: false
---

This vignette walks through a complete whole-slide inference workflow using `petrographer`. It mirrors the experience of a user who has a scanned thin section, wants to run the pretrained detector with SAHI slicing, and then review both the visual overlays and quantitative statistics.

## Load the Toolkit

The package ships with all high-level helpers. Load it alongside a few tidy utilities you may want for inspection.

```{r setup}
library(petrographer)
library(dplyr)
library(fs)
```

## Point to Your Whole-Slide Image

Set the path to the slide you want to analyse. Replace the placeholder with your `.svs`, `.tif`, or other supported image file.

```{r whole-slide-path, eval=FALSE}
wsi_path <- system.file("extdata", "test_image.jpg", package = "petrographer")
file_exists(wsi_path)
```

Swap in your `.svs`, `.tif`, or other supported whole-slide file once you are ready to run the full workflow.

## Cache a Pretrained Model

Pretrained weights are hosted on the petrographer hub. Install (or refresh) the pin you need, then instantiate the R-side wrapper. Replace `model_id` with whichever detector matches your slide.

```{r load-model, eval=FALSE}
model <- from_pretrained('inclusions_shell')
```

`pg_install_pretrained()` downloads the pin once and reuses the cache next time. The returned `PetrographyModel` retains the manifest so you can inspect provenance or training metadata as needed.

## Run SAHI Inference with Custom Strides

Whole-slide images are large, so enabling SAHI slicing keeps detections accurate while fitting within GPU or CPU memory. Control the stride through `slice_size` and `overlap`.

```{r predict-slide, eval=FALSE}
output_dir <- path("results", path_file(path_ext_remove(wsi_path)))

predictions <- predict_image(
  image_path = wsi_path,
  model = model,
  use_slicing = FALSE,
  slice_size = 1024,  # pixels per tile; shrink for tighter memory budgets
  overlap = 0.2,     # tile overlap to smooth boundaries
  output_dir = output_dir,
  save_visualizations = TRUE
)

predictions
```

The tibble contains one row per detected grain, complete with geometric measurements and morphology descriptors ready for downstream analysis.

## Visualise the Overlay and Exports

SAHI writes annotated overlays alongside the raw detections. Locate the generated assets and preview them in-line.

```{r visualise-results, eval=FALSE}
dir_ls(output_dir)

viz_path <- dir_ls(output_dir, glob = "*_prediction.png", fail = FALSE)
if (length(viz_path)) {
  image_read(viz_path[[1]])
}
```

You can also hand the exported visual to image viewers or documentation pipelines — each file is saved with the original slide name plus `_prediction.png`.

## Summarise Detection Statistics

With the predictions tibble in hand, compute per-image summaries or overall population metrics to quantify mineralogy at scale.

```{r summarise-stats, eval=FALSE}
per_image <- summarize_by_image(predictions)
per_image

population <- get_population_stats(predictions)
population
```

These summaries surface object counts, size distributions, and shape metrics that support QA checks or comparisons between slides.

## Next Steps

- Adjust `slice_size` and `overlap` to balance runtime and fidelity for extremely large slides.
- Pipe `predictions` into custom tidyverse workflows for further filtering, plotting, or reporting.
- Use `predict_images()` to batch-process entire slide directories once you are satisfied with the single-slide results.
