---
title: "Train Detection Library Models"
format:
  html:
    toc: true
params:
  dataset_id: "inclusions_shell"
  data_dir: "data/processed/inclusions_shell"
  dataset_description: "Detection training dataset on mixed shell and inclusion"
  model_id_prefix: "inclusions_shell"
  variants: ["nano", "small", "medium", "large"]
  epochs: 120
  batch_size: 8
  time_hours: 3
  validate_every: 2
  early_stopping_patience: null
  confidence: 0.5
  device: "mps"
  slice_size: 1024
  overlap: 0.2
  max_dets: 300
  test_image_index: 1
  hpc_host: "hpg"
  hpc_base_dir: "/blue/nicolas.gauthier/nicolas.gauthier/petrographer"
  publish_to_hub: false
  hub_board_url: null
---

Operational notebook for training multiple **detection** variants for the model
library with one shared configuration.

This notebook:

- trains each variant in sequence
- continues past failures and records them
- runs a local load/predict check for successful runs
- computes SAHI validation metrics for successful runs
- optionally republishes successful runs to a shared hub board

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(fs)
library(here)
library(pins)
library(petrographer)
library(hipergator)

variants <- unlist(params$variants)
stopifnot(length(variants) > 0)
stopifnot(all(variants %in% c("nano", "small", "medium", "large")))

hpg_configure(
  host = params$hpc_host,
  base_dir = params$hpc_base_dir
)

data_dir <- here(params$data_dir)
out_dir <- here("results", "ops_detection_library")
dir_create(out_dir)

pin_dataset(
  data_dir = data_dir,
  dataset_id = params$dataset_id,
  metadata = list(
    description = params$dataset_description
  )
)

dataset_board <- petrographer:::.get_dataset_board()
dataset_path <- get_dataset_path(params$dataset_id, board = "local")
dataset_meta <- pins::pin_meta(dataset_board, params$dataset_id)

valid_images <- dir_ls(path(data_dir, "valid"), glob = "*.jpg")
if (!length(valid_images)) {
  valid_images <- dir_ls(path(data_dir, "valid"), glob = "*.png")
}
stopifnot(length(valid_images) > 0)

test_image <- valid_images[[params$test_image_index]]

hub_board <- if (isTRUE(params$publish_to_hub)) {
  stopifnot(!is.null(params$hub_board_url), nzchar(params$hub_board_url))
  board_url(params$hub_board_url)
} else {
  NULL
}
```

## Submit Variants

```{r submit-library}
handles <- vector("list", length(variants))

for (i in seq_along(variants)) {
  variant <- variants[[i]]
  model_id <- paste(params$model_id_prefix, variant, sep = "_")

  config <- petrographer:::prepare_training_config(
    data_dir = dataset_path,
    dataset_id = params$dataset_id,
    dataset_version = dataset_meta$version,
    model_id = model_id,
    model_variant = variant,
    resolution = NULL,
    epochs = params$epochs,
    grad_accum_steps = NA,
    learning_rate = NULL,
    device = "cuda",
    batch_size = params$batch_size,
    use_amp = TRUE,
    amp_dtype = "bf16",
    gradient_checkpointing = FALSE,
    num_workers = NULL,
    time_hours = params$time_hours,
    validate_every = params$validate_every,
    early_stopping_patience = params$early_stopping_patience
  )

  handles[[i]] <- tryCatch(
    petrographer:::submit_hpc_training(config),
    error = function(e) list(
      config = config,
      submit_error = conditionMessage(e),
      submitted_at = Sys.time()
    )
  )
}
```

## Wait, Collect, and Validate

```{r train-library}
run_records <- vector("list", length(variants))

for (i in seq_along(variants)) {
  variant <- variants[[i]]
  handle <- handles[[i]]
  model_id <- handle$config$model_id

  if (!is.null(handle$submit_error)) {
    run_records[[i]] <- tibble::tibble(
      variant = variant,
      model_id = model_id,
      status = "failed",
      detections = NA_integer_,
      mean_confidence = NA_real_,
      ap = NA_real_,
      ap50 = NA_real_,
      ar = NA_real_,
      published = FALSE,
      error = handle$submit_error
    )
    next
  }

  run_records[[i]] <- tryCatch({
    handle <- petrographer:::wait_hpc_training(handle)
    model_dir <- petrographer:::collect_hpc_training(handle)

    completed_at <- if (!is.null(handle$completed_at)) handle$completed_at else Sys.time()
    duration_mins <- round(
      as.numeric(difftime(completed_at, handle$submitted_at, units = "mins")),
      1
    )

    trained_model_id <- petrographer:::finalize_trained_model(
      model_dir = model_dir,
      config = handle$config,
      duration_mins = duration_mins
    )

    model <- from_pretrained(
      trained_model_id,
      board = "local",
      device = params$device,
      confidence = params$confidence
    )

    sanity_result <- predict_image(
      image_path = test_image,
      model = model,
      use_slicing = TRUE,
      slice_size = params$slice_size,
      overlap = params$overlap,
      output_dir = path(out_dir, trained_model_id)
    )

    training_eval <- evaluate_training(
      model_id = trained_model_id,
      output_dir = path(out_dir, trained_model_id, "training_eval")
    )

    sahi_eval <- evaluate_model_sahi(
      model = model,
      annotation_json = path(data_dir, "valid/_annotations.coco.json"),
      image_dir = path(data_dir, "valid"),
      use_slicing = TRUE,
      slice_size = params$slice_size,
      overlap = params$overlap,
      max_dets = params$max_dets,
      save_predictions = path(out_dir, trained_model_id, "sahi_predictions.json")
    )

    # Per-class SAHI metrics: write per variant so the post-loop chunk can read
    # them back without holding all sahi_eval objects in memory.
    if (!is.null(sahi_eval$per_class) && nrow(sahi_eval$per_class) > 0) {
      readr::write_csv(
        sahi_eval$per_class,
        path(out_dir, trained_model_id, "per_class_metrics.csv")
      )
    }

    # Confusion matrix at IoU >= 0.5. Greedy matching; see ?confusion_matrix.
    cm <- confusion_matrix(
      predictions = sahi_eval$predictions,
      annotation_json = path(data_dir, "valid/_annotations.coco.json"),
      iou_threshold = 0.5,
      score_threshold = params$confidence
    )
    readr::write_csv(
      cm$long,
      path(out_dir, trained_model_id, "confusion_matrix.csv")
    )

    published <- FALSE
    if (isTRUE(params$publish_to_hub)) {
      local_board <- pins::board_folder(here(".petrographer/models"), versioned = TRUE)
      model_files <- pins::pin_download(local_board, trained_model_id)
      model_dir <- fs::path_dir(model_files[[1]])
      pin_model(
        model_dir = model_dir,
        model_id = trained_model_id,
        board = hub_board,
        metadata = list(notes = "Published from train_detection_library.qmd")
      )
      published <- TRUE
    }

    summary_lookup <- stats::setNames(sahi_eval$summary$value, sahi_eval$summary$metric)

    tibble::tibble(
      variant = variant,
      model_id = trained_model_id,
      status = "success",
      detections = nrow(sanity_result),
      mean_confidence = if (nrow(sanity_result) > 0) mean(sanity_result$confidence) else NA_real_,
      ap = unname(summary_lookup["AP"]),
      ap50 = unname(summary_lookup["AP50"]),
      ar = unname(summary_lookup[paste0("AR@", params$max_dets)]),
      published = published,
      error = NA_character_
    )
  }, error = function(e) {
    tibble::tibble(
      variant = variant,
      model_id = model_id,
      status = "failed",
      detections = NA_integer_,
      mean_confidence = NA_real_,
      ap = NA_real_,
      ap50 = NA_real_,
      ar = NA_real_,
      published = FALSE,
      error = conditionMessage(e)
    )
  })
}

library_runs <- bind_rows(run_records)
readr::write_csv(library_runs, path(out_dir, "detection_library_runs.csv"))
library_runs
```

## Compare Successful Variants

```{r compare}
library_runs |>
  filter(status == "success") |>
  arrange(desc(ap))
```

```{r compare-plot, eval=sum(library_runs$status == "success") > 0}
library_runs |>
  filter(status == "success") |>
  mutate(variant = factor(variant, levels = variants)) |>
  pivot_longer(c(ap, ap50, ar), names_to = "metric", values_to = "value") |>
  ggplot(aes(variant, value, fill = metric)) +
  geom_col(position = "dodge") +
  labs(
    title = "Detection variant comparison",
    y = "metric value"
  ) +
  theme_minimal()
```

## Training Curves

Read back the per-variant metrics that `evaluate_training()` wrote during the
loop (one `training_eval/` directory per variant). The faceted plots below let
you check whether each variant plateaued or was still improving when training
stopped — if validation mAP is still rising at the last epoch, you should
train longer; if val loss is rising while train loss falls, you're overfitting.

```{r load-curves, eval=sum(library_runs$status == "success") > 0}
successful <- library_runs |> filter(status == "success")

read_variant_curves <- function(variant, model_id) {
  eval_dir <- path(out_dir, model_id, "training_eval")
  train_path <- path(eval_dir, "training_metrics.csv")
  val_path   <- path(eval_dir, "validation_metrics.csv")

  train_tbl <- if (file_exists(train_path)) {
    readr::read_csv(train_path, show_col_types = FALSE) |>
      mutate(variant = variant, split = "train")
  } else {
    tibble::tibble(variant = variant, split = "train", epoch = integer())
  }
  val_tbl <- if (file_exists(val_path)) {
    readr::read_csv(val_path, show_col_types = FALSE) |>
      mutate(variant = variant, split = "val")
  } else {
    tibble::tibble(variant = variant, split = "val", epoch = integer())
  }
  list(train = train_tbl, val = val_tbl)
}

all_curves <- purrr::map2(successful$variant, successful$model_id, read_variant_curves)
training_curves   <- bind_rows(purrr::map(all_curves, "train")) |>
  mutate(variant = factor(variant, levels = variants))
validation_curves <- bind_rows(purrr::map(all_curves, "val")) |>
  mutate(variant = factor(variant, levels = variants))
```

### Validation mAP over epochs

Primary convergence signal. Flat line at the end = trained enough. Still
climbing = train longer.

```{r val-map-plot, eval=nrow(validation_curves) > 0 && "mAP_50_95" %in% names(validation_curves)}
validation_curves |>
  select(variant, epoch, any_of(c("mAP_50_95", "mAP_50"))) |>
  pivot_longer(-c(variant, epoch), names_to = "metric", values_to = "value") |>
  filter(!is.na(value)) |>
  ggplot(aes(epoch, value, color = variant)) +
  geom_line() +
  geom_point(size = 0.8, alpha = 0.6) +
  facet_wrap(~ metric, scales = "free_y") +
  labs(
    title = "Validation mAP by variant",
    y = "mAP",
    x = "epoch"
  ) +
  theme_minimal()
```

### Train vs val loss

Overfitting check. Train loss should fall; val loss should fall then plateau.
If val loss turns upward while train loss keeps falling, the model is
memorizing — earlier stopping (or more regularization) would help.

```{r loss-plot, eval=nrow(training_curves) > 0 && "loss" %in% names(training_curves)}
train_loss <- training_curves |>
  select(variant, epoch, loss) |>
  filter(!is.na(loss)) |>
  mutate(split = "train")
val_loss <- if ("loss" %in% names(validation_curves)) {
  validation_curves |>
    select(variant, epoch, loss) |>
    filter(!is.na(loss)) |>
    mutate(split = "val")
} else {
  tibble::tibble(variant = character(), epoch = integer(), loss = numeric(), split = character())
}

bind_rows(train_loss, val_loss) |>
  ggplot(aes(epoch, loss, color = split)) +
  geom_line() +
  facet_wrap(~ variant, scales = 'free_y') +
  labs(
    title = "Train vs validation loss",
    y = "loss",
    x = "epoch"
  ) +
  theme_minimal()
```

### Final-epoch snapshot

Quick numeric answer to "did I train long enough?" — if the last few epochs
show the metrics still climbing by >0.5% per epoch, keep going.

```{r final-epochs, eval=nrow(validation_curves) > 0}
validation_curves |>
  group_by(variant) |>
  slice_tail(n = 5) |>
  select(variant, epoch, any_of(c("mAP_50_95", "mAP_50", "loss"))) |>
  arrange(variant, epoch)
```

## Per-Class Analysis

Multi-class datasets: which classes is each variant good/bad at? Three views:

1. **Per-class training curves** — how each class's mAP evolved over epochs (from
   rfdetr's `--log-per-class-metrics` written to `validation_classwise.csv`).
2. **Per-class SAHI eval** — final AP / AP50 / AR per class on the validation
   set (written by `evaluate_model_sahi()` in the main loop).
3. **Confusion matrix** — where class mistakes happen (written by
   `confusion_matrix()` in the main loop). Matched by IoU >= 0.5; unmatched
   predictions go to the `background` row, unmatched GT to the `background`
   column.

### Per-class training curves

Per-class mAP trajectories. Classes that plateau early while others keep
climbing usually mean the plateaued class is easy and the climbing one is
still learning. Classes that diverge (one climbs, another drops) signal
class imbalance pulling the optimizer.

```{r per-class-curves, eval=sum(library_runs$status == "success") > 0}
per_class_curves <- purrr::map_dfr(successful$model_id, function(mid) {
  cw_path <- path(out_dir, mid, "training_eval", "validation_classwise.csv")
  if (!file_exists(cw_path)) return(NULL)
  readr::read_csv(cw_path, show_col_types = FALSE) |>
    mutate(model_id = mid,
           variant = successful$variant[successful$model_id == mid])
})

if (nrow(per_class_curves) > 0 && "map_50_95" %in% names(per_class_curves)) {
  per_class_curves |>
    mutate(variant = factor(variant, levels = variants)) |>
    ggplot(aes(epoch, map_50_95, color = class_name)) +
    geom_line(alpha = 0.8) +
    geom_point(size = 0.7, alpha = 0.6) +
    facet_wrap(~ variant) +
    labs(
      title = "Per-class validation mAP@[.5:.95] by variant",
      y = "mAP", x = "epoch", color = "class"
    ) +
    theme_minimal()
}
```

### Per-class SAHI eval

Final per-class metrics on the validation set, evaluated with SAHI slicing
at `slice_size = {r params$slice_size}`. Compare across variants at a glance.

```{r per-class-sahi, eval=sum(library_runs$status == "success") > 0}
per_class_sahi <- purrr::map_dfr(successful$model_id, function(mid) {
  pc_path <- path(out_dir, mid, "per_class_metrics.csv")
  if (!file_exists(pc_path)) return(NULL)
  readr::read_csv(pc_path, show_col_types = FALSE) |>
    mutate(model_id = mid,
           variant = successful$variant[successful$model_id == mid])
})
per_class_sahi
```

```{r per-class-sahi-plot, eval=exists("per_class_sahi") && nrow(per_class_sahi) > 0}
per_class_sahi |>
  mutate(variant = factor(variant, levels = variants)) |>
  pivot_longer(c(AP, AP50, AR), names_to = "metric", values_to = "value") |>
  ggplot(aes(class_name, value, fill = variant)) +
  geom_col(position = "dodge") +
  facet_wrap(~ metric, scales = "free_y") +
  labs(
    title = "Per-class SAHI eval metrics by variant",
    x = "class", y = "value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

### Confusion matrices

Rows = ground-truth class, columns = predicted class. The `background` row
counts false positives (predictions with no GT match); the `background`
column counts false negatives (GT boxes nothing matched). Diagonal = correct
class, off-diagonal = class confusion at IoU >= 0.5.

```{r confusion-matrices, eval=sum(library_runs$status == "success") > 0, fig.height=6}
confusion_long <- purrr::map_dfr(successful$model_id, function(mid) {
  cm_path <- path(out_dir, mid, "confusion_matrix.csv")
  if (!file_exists(cm_path)) return(NULL)
  readr::read_csv(cm_path, show_col_types = FALSE) |>
    mutate(model_id = mid,
           variant = successful$variant[successful$model_id == mid])
})

if (nrow(confusion_long) > 0) {
  confusion_long |>
    mutate(
      variant = factor(variant, levels = variants),
      gt_label = factor(gt_label, levels = unique(confusion_long$gt_label)),
      pred_label = factor(pred_label, levels = levels(gt_label))
    ) |>
    ggplot(aes(pred_label, gt_label, fill = count)) +
    geom_tile() +
    geom_text(aes(label = count), size = 3) +
    facet_wrap(~ variant) +
    scale_fill_gradient(low = "white", high = "steelblue", trans = "log1p") +
    labs(
      title = "Confusion matrices (IoU >= 0.5)",
      x = "predicted class",
      y = "true class"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}
```

## Failures

```{r failures}
library_runs |>
  filter(status == "failed") |>
  select(variant, model_id, error)
```
