---
title: "Train Segmentation Library Models"
format:
  html:
    toc: true
params:
  dataset_id: "inclusions_shell"
  data_dir: "data/processed/inclusions_shell_sliced"
  dataset_description: "Segmentation training dataset"
  model_id_prefix: "inclusions_shell"
  variants: ["seg_nano", "seg_small", "seg_medium", "seg_large"]
  epochs: 120
  batch_size: 8
  time_hours: 8
  validate_every: 2
  early_stopping_patience: null
  confidence: 0.5
  device: "mps"
  test_image_index: 1
  hpc_host: "hpg"
  hpc_base_dir: null
  publish_to_hub: false
  hub_board_url: null
---

Operational notebook for training multiple **segmentation** variants for the
model library with one shared configuration.

This notebook:

- trains each variant in sequence
- continues past failures and records them
- runs a local load/predict check for successful runs
- exports morphology summaries for successful runs
- optionally republishes successful runs to a shared hub board

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(fs)
library(here)
library(pins)
library(petrographer)
library(hipergator)

variants <- unlist(params$variants)
stopifnot(length(variants) > 0)
stopifnot(all(variants %in% c("seg_nano", "seg_small", "seg_medium", "seg_large",
                              "seg_xlarge", "seg_2xlarge", "seg_preview")))

hpg_configure(
  host = params$hpc_host,
  base_dir = params$hpc_base_dir
)

data_dir <- here(params$data_dir)
out_dir <- here("results", "ops_segmentation_library")
dir_create(out_dir)

pin_dataset(
  data_dir = data_dir,
  dataset_id = params$dataset_id,
  metadata = list(
    description = params$dataset_description
  )
)

dataset_board <- petrographer:::.get_dataset_board()
dataset_path <- get_dataset_path(params$dataset_id, board = "local")
dataset_meta <- pins::pin_meta(dataset_board, params$dataset_id)

valid_images <- dir_ls(path(data_dir, "valid"), glob = "*.jpg")
if (!length(valid_images)) {
  valid_images <- dir_ls(path(data_dir, "valid"), glob = "*.png")
}
stopifnot(length(valid_images) > 0)

test_image <- valid_images[[params$test_image_index]]

hub_board <- if (isTRUE(params$publish_to_hub)) {
  stopifnot(!is.null(params$hub_board_url), nzchar(params$hub_board_url))
  board_url(params$hub_board_url)
} else {
  NULL
}
```

## Submit Variants

```{r submit-library}
handles <- vector("list", length(variants))

for (i in seq_along(variants)) {
  variant <- variants[[i]]
  model_id <- paste(params$model_id_prefix, variant, sep = "_")

  config <- petrographer:::prepare_training_config(
    data_dir = dataset_path,
    dataset_id = params$dataset_id,
    dataset_version = dataset_meta$version,
    model_id = model_id,
    model_variant = variant,
    resolution = NULL,
    epochs = params$epochs,
    grad_accum_steps = NA,
    learning_rate = NULL,
    device = "cuda",
    batch_size = params$batch_size,
    use_amp = TRUE,
    amp_dtype = "bf16",
    gradient_checkpointing = FALSE,
    num_workers = NULL,
    time_hours = params$time_hours,
    validate_every = params$validate_every,
    early_stopping_patience = params$early_stopping_patience
  )

  handles[[i]] <- tryCatch(
    petrographer:::submit_hpc_training(config),
    error = function(e) list(
      config = config,
      submit_error = conditionMessage(e),
      submitted_at = Sys.time()
    )
  )
}
```

## Wait, Collect, and Validate

```{r train-library}
run_records <- vector("list", length(variants))

for (i in seq_along(variants)) {
  variant <- variants[[i]]
  handle <- handles[[i]]
  model_id <- handle$config$model_id

  if (!is.null(handle$submit_error)) {
    run_records[[i]] <- tibble::tibble(
      variant = variant,
      model_id = model_id,
      status = "failed",
      detections = NA_integer_,
      mean_confidence = NA_real_,
      mean_area = NA_real_,
      total_objects = NA_real_,
      published = FALSE,
      error = handle$submit_error
    )
    next
  }

  run_records[[i]] <- tryCatch({
    handle <- petrographer:::wait_hpc_training(handle)
    model_dir <- petrographer:::collect_hpc_training(handle)

    completed_at <- if (!is.null(handle$completed_at)) handle$completed_at else Sys.time()
    duration_mins <- round(
      as.numeric(difftime(completed_at, handle$submitted_at, units = "mins")),
      1
    )

    trained_model_id <- petrographer:::finalize_trained_model(
      model_dir = model_dir,
      config = handle$config,
      duration_mins = duration_mins
    )

    model <- from_pretrained(
      trained_model_id,
      board = "local",
      device = params$device,
      confidence = params$confidence
    )

    sanity_result <- predict_image(
      image_path = test_image,
      model = model,
      use_slicing = FALSE,
      output_dir = path(out_dir, trained_model_id)
    )

    training_eval <- evaluate_training(
      model_id = trained_model_id,
      output_dir = path(out_dir, trained_model_id, "training_eval")
    )

    seg_batch <- analyze_segmentation_dir(
      input_dir = path(data_dir, "valid"),
      model = model,
      output_dir = path(out_dir, trained_model_id, "segmentation_batch")
    )

    published <- FALSE
    if (isTRUE(params$publish_to_hub)) {
      local_board <- pins::board_folder(here(".petrographer/models"), versioned = TRUE)
      model_files <- pins::pin_download(local_board, trained_model_id)
      model_dir <- fs::path_dir(model_files[[1]])
      pin_model(
        model_dir = model_dir,
        model_id = trained_model_id,
        board = hub_board,
        metadata = list(notes = "Published from train_segmentation_library.qmd")
      )
      published <- TRUE
    }

    tibble::tibble(
      variant = variant,
      model_id = trained_model_id,
      status = "success",
      detections = nrow(sanity_result),
      mean_confidence = if (nrow(sanity_result) > 0) mean(sanity_result$confidence) else NA_real_,
      mean_area = seg_batch$population_stats$mean_area %||% NA_real_,
      total_objects = seg_batch$population_stats$total_objects %||% NA_real_,
      published = published,
      error = NA_character_
    )
  }, error = function(e) {
    tibble::tibble(
      variant = variant,
      model_id = model_id,
      status = "failed",
      detections = NA_integer_,
      mean_confidence = NA_real_,
      mean_area = NA_real_,
      total_objects = NA_real_,
      published = FALSE,
      error = conditionMessage(e)
    )
  })
}

library_runs <- bind_rows(run_records)
readr::write_csv(library_runs, path(out_dir, "segmentation_library_runs.csv"))
library_runs
```

## Compare Successful Variants

```{r compare}
library_runs |>
  filter(status == "success") |>
  arrange(desc(total_objects))
```

```{r compare-plot, eval=sum(library_runs$status == "success") > 0}
library_runs |>
  filter(status == "success") |>
  pivot_longer(c(mean_confidence, mean_area, total_objects), names_to = "metric", values_to = "value") |>
  ggplot(aes(variant, value, fill = metric)) +
  geom_col(position = "dodge") +
  labs(
    title = "Segmentation variant comparison",
    y = "metric value"
  ) +
  theme_minimal()
```

## Training Curves

Read back the per-variant metrics that `evaluate_training()` wrote during the
loop (one `training_eval/` directory per variant). Check whether each variant
plateaued or was still improving when training stopped.

```{r load-curves, eval=sum(library_runs$status == "success") > 0}
successful <- library_runs |> filter(status == "success")

read_variant_curves <- function(variant, model_id) {
  eval_dir <- path(out_dir, model_id, "training_eval")
  train_path <- path(eval_dir, "training_metrics.csv")
  val_path   <- path(eval_dir, "validation_metrics.csv")
  cw_path    <- path(eval_dir, "validation_classwise.csv")

  train_tbl <- if (file_exists(train_path)) {
    readr::read_csv(train_path, show_col_types = FALSE) |>
      mutate(variant = variant, split = "train")
  } else {
    tibble::tibble(variant = variant, split = "train", epoch = integer())
  }
  val_tbl <- if (file_exists(val_path)) {
    readr::read_csv(val_path, show_col_types = FALSE) |>
      mutate(variant = variant, split = "val")
  } else {
    tibble::tibble(variant = variant, split = "val", epoch = integer())
  }
  cw_tbl <- if (file_exists(cw_path)) {
    readr::read_csv(cw_path, show_col_types = FALSE) |>
      mutate(variant = variant)
  } else {
    tibble::tibble(variant = variant, epoch = integer(), class_name = character())
  }
  list(train = train_tbl, val = val_tbl, classwise = cw_tbl)
}

all_curves <- purrr::map2(successful$variant, successful$model_id, read_variant_curves)
training_curves   <- bind_rows(purrr::map(all_curves, "train")) |>
  mutate(variant = factor(variant, levels = variants))
validation_curves <- bind_rows(purrr::map(all_curves, "val")) |>
  mutate(variant = factor(variant, levels = variants))
classwise_curves  <- bind_rows(purrr::map(all_curves, "classwise")) |>
  mutate(variant = factor(variant, levels = variants))
```

### Validation mAP over epochs

Primary convergence signal. Flat = trained enough. Still climbing = train longer.

```{r val-map-plot, eval=nrow(validation_curves) > 0 && "mAP_50_95" %in% names(validation_curves)}
validation_curves |>
  select(variant, epoch, any_of(c("mAP_50_95", "mAP_50"))) |>
  pivot_longer(-c(variant, epoch), names_to = "metric", values_to = "value") |>
  filter(!is.na(value)) |>
  ggplot(aes(epoch, value, color = variant)) +
  geom_line() +
  geom_point(size = 0.8, alpha = 0.6) +
  facet_wrap(~ metric, scales = "free_y") +
  labs(title = "Validation mAP by variant", y = "mAP", x = "epoch") +
  theme_minimal()
```

### Train vs val loss

Overfitting check. If val loss turns upward while train loss keeps falling,
the model is memorizing.

```{r loss-plot, eval=nrow(training_curves) > 0 && "loss" %in% names(training_curves)}
train_loss <- training_curves |>
  select(variant, epoch, loss) |>
  filter(!is.na(loss)) |>
  mutate(split = "train")
val_loss <- if ("loss" %in% names(validation_curves)) {
  validation_curves |>
    select(variant, epoch, loss) |>
    filter(!is.na(loss)) |>
    mutate(split = "val")
} else {
  tibble::tibble(variant = character(), epoch = integer(), loss = numeric(), split = character())
}

bind_rows(train_loss, val_loss) |>
  ggplot(aes(epoch, loss, color = split)) +
  geom_line() +
  facet_wrap(~ variant, scales = "free_y") +
  labs(title = "Train vs validation loss", y = "loss", x = "epoch") +
  theme_minimal()
```

### Per-class validation mAP

For multi-class datasets: which classes is each variant good/bad at? Classes
that plateau early while others keep climbing usually mean the plateaued
class is easy and the climbing one is still learning. Large gaps between
classes are worth flagging — could be class imbalance or annotation quality.

```{r per-class-curves, eval=nrow(classwise_curves) > 0 && "map_50_95" %in% names(classwise_curves)}
classwise_curves |>
  filter(!is.na(map_50_95)) |>
  ggplot(aes(epoch, map_50_95, color = class_name)) +
  geom_line(alpha = 0.8) +
  geom_point(size = 0.7, alpha = 0.6) +
  facet_wrap(~ variant) +
  labs(
    title = "Per-class validation mAP@[.5:.95] by variant",
    y = "mAP", x = "epoch", color = "class"
  ) +
  theme_minimal()
```

### Final-epoch snapshot

```{r final-epochs, eval=nrow(validation_curves) > 0}
validation_curves |>
  group_by(variant) |>
  slice_tail(n = 5) |>
  select(variant, epoch, any_of(c("mAP_50_95", "mAP_50", "loss"))) |>
  arrange(variant, epoch)
```

## Failures

```{r failures}
library_runs |>
  filter(status == "failed") |>
  select(variant, model_id, error)
```
