---
title: "Train Detection Library Models"
format:
  html:
    toc: true
params:
  dataset_id: "inclusions"
  data_dir: "data/processed/inclusions_sliced"
  dataset_description: "Detection training dataset on sliced inclusions"
  model_id_prefix: "inclusions"
  variants: ["nano", "small", "medium", "large"]
  epochs: 100
  batch_size: "auto"
  time_hours: 4
  validate_every: 2
  early_stopping_patience: 10
  confidence: 0.5
  device: "cpu"
  slice_size: 640
  overlap: 0.2
  max_dets: 300
  test_image_index: 1
  hpc_host: "hpg"
  hpc_base_dir: "/blue/nicolas.gauthier/nicolas.gauthier/petrographer"
  publish_to_hub: false
  hub_board_url: null
---

Operational notebook for training multiple **detection** variants for the model
library with one shared configuration.

This notebook:

- trains each variant in sequence
- continues past failures and records them
- runs a local load/predict check for successful runs
- computes SAHI validation metrics for successful runs
- optionally republishes successful runs to a shared hub board

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(fs)
library(here)
library(pins)
library(petrographer)
library(hipergator)

variants <- unlist(params$variants)
stopifnot(length(variants) > 0)
stopifnot(all(variants %in% c("nano", "small", "medium", "large")))

hpg_configure(
  host = params$hpc_host,
  base_dir = params$hpc_base_dir
)

data_dir <- here(params$data_dir)
out_dir <- here("results", "ops_detection_library")
dir_create(out_dir)

pin_dataset(
  data_dir = data_dir,
  dataset_id = params$dataset_id,
  metadata = list(
    description = params$dataset_description
  )
)

dataset_board <- petrographer:::.get_dataset_board()
dataset_path <- get_dataset_path(params$dataset_id, board = "local")
dataset_meta <- pins::pin_meta(dataset_board, params$dataset_id)

valid_images <- dir_ls(path(data_dir, "valid"), glob = "*.jpg")
if (!length(valid_images)) {
  valid_images <- dir_ls(path(data_dir, "valid"), glob = "*.png")
}
stopifnot(length(valid_images) > 0)

test_image <- valid_images[[params$test_image_index]]

hub_board <- if (isTRUE(params$publish_to_hub)) {
  stopifnot(!is.null(params$hub_board_url), nzchar(params$hub_board_url))
  board_url(params$hub_board_url)
} else {
  NULL
}
```

## Submit Variants

```{r submit-library}
handles <- vector("list", length(variants))

for (i in seq_along(variants)) {
  variant <- variants[[i]]
  model_id <- paste(params$model_id_prefix, variant, sep = "_")

  config <- petrographer:::prepare_training_config(
    data_dir = dataset_path,
    dataset_id = params$dataset_id,
    dataset_version = dataset_meta$version,
    model_id = model_id,
    model_variant = variant,
    resolution = NULL,
    epochs = params$epochs,
    grad_accum_steps = NA,
    learning_rate = NULL,
    device = "cuda",
    batch_size = params$batch_size,
    use_amp = TRUE,
    amp_dtype = "bf16",
    gradient_checkpointing = FALSE,
    num_workers = NULL,
    time_hours = params$time_hours,
    validate_every = params$validate_every,
    early_stopping_patience = params$early_stopping_patience
  )

  handles[[i]] <- tryCatch(
    petrographer:::submit_hpc_training(config),
    error = function(e) list(
      config = config,
      submit_error = conditionMessage(e),
      submitted_at = Sys.time()
    )
  )
}
```

## Wait, Collect, and Validate

```{r train-library}
run_records <- vector("list", length(variants))

for (i in seq_along(variants)) {
  variant <- variants[[i]]
  handle <- handles[[i]]
  model_id <- handle$config$model_id

  if (!is.null(handle$submit_error)) {
    run_records[[i]] <- tibble::tibble(
      variant = variant,
      model_id = model_id,
      status = "failed",
      detections = NA_integer_,
      mean_confidence = NA_real_,
      ap = NA_real_,
      ap50 = NA_real_,
      ar = NA_real_,
      published = FALSE,
      error = handle$submit_error
    )
    next
  }

  run_records[[i]] <- tryCatch({
    handle <- petrographer:::wait_hpc_training(handle)
    model_dir <- petrographer:::collect_hpc_training(handle)

    completed_at <- if (!is.null(handle$completed_at)) handle$completed_at else Sys.time()
    duration_mins <- round(
      as.numeric(difftime(completed_at, handle$submitted_at, units = "mins")),
      1
    )

    trained_model_id <- petrographer:::finalize_trained_model(
      model_dir = model_dir,
      config = handle$config,
      duration_mins = duration_mins
    )

    model <- from_pretrained(
      trained_model_id,
      board = "local",
      device = params$device,
      confidence = params$confidence
    )

    sanity_result <- predict_image(
      image_path = test_image,
      model = model,
      use_slicing = TRUE,
      slice_size = params$slice_size,
      overlap = params$overlap,
      output_dir = path(out_dir, trained_model_id)
    )

    training_eval <- evaluate_training(
      model_id = trained_model_id,
      output_dir = path(out_dir, trained_model_id, "training_eval")
    )

    sahi_eval <- evaluate_model_sahi(
      model = model,
      annotation_json = path(data_dir, "valid/_annotations.coco.json"),
      image_dir = path(data_dir, "valid"),
      use_slicing = TRUE,
      slice_size = params$slice_size,
      overlap = params$overlap,
      max_dets = params$max_dets,
      save_predictions = path(out_dir, trained_model_id, "sahi_predictions.json")
    )

    published <- FALSE
    if (isTRUE(params$publish_to_hub)) {
      local_board <- pins::board_folder(here(".petrographer/models"), versioned = TRUE)
      model_files <- pins::pin_download(local_board, trained_model_id)
      model_dir <- fs::path_dir(model_files[[1]])
      pin_model(
        model_dir = model_dir,
        model_id = trained_model_id,
        board = hub_board,
        metadata = list(notes = "Published from train_detection_library.qmd")
      )
      published <- TRUE
    }

    summary_lookup <- stats::setNames(sahi_eval$summary$value, sahi_eval$summary$metric)

    tibble::tibble(
      variant = variant,
      model_id = trained_model_id,
      status = "success",
      detections = nrow(sanity_result),
      mean_confidence = if (nrow(sanity_result) > 0) mean(sanity_result$confidence) else NA_real_,
      ap = unname(summary_lookup["AP"]),
      ap50 = unname(summary_lookup["AP50"]),
      ar = unname(summary_lookup[paste0("AR@", params$max_dets)]),
      published = published,
      error = NA_character_
    )
  }, error = function(e) {
    tibble::tibble(
      variant = variant,
      model_id = model_id,
      status = "failed",
      detections = NA_integer_,
      mean_confidence = NA_real_,
      ap = NA_real_,
      ap50 = NA_real_,
      ar = NA_real_,
      published = FALSE,
      error = conditionMessage(e)
    )
  })
}

library_runs <- bind_rows(run_records)
readr::write_csv(library_runs, path(out_dir, "detection_library_runs.csv"))
library_runs
```

## Compare Successful Variants

```{r compare}
library_runs |>
  filter(status == "success") |>
  arrange(desc(ap))
```

```{r compare-plot, eval=sum(library_runs$status == "success") > 0}
library_runs |>
  filter(status == "success") |>
  pivot_longer(c(ap, ap50, ar), names_to = "metric", values_to = "value") |>
  ggplot(aes(variant, value, fill = metric)) +
  geom_col(position = "dodge") +
  labs(
    title = "Detection variant comparison",
    y = "metric value"
  ) +
  theme_minimal()
```

## Failures

```{r failures}
library_runs |>
  filter(status == "failed") |>
  select(variant, model_id, error)
```
