---
title: "Petrographer Model Zoo with pins"
author: "petrographer team"
format:
  html:
    code-fold: true
    toc: true
editor: visual
---

> **Reference notebook, not supported user docs**
>
> This notebook is kept as maintainer-facing reference for board management and
> publishing workflows. It is useful for mining patterns and commands, but it is
> not treated as a polished public tutorial.

```{r}
#| label: setup
#| message: false
#| warning: false
library(petrographer)
library(here)
if (requireNamespace("pins", quietly = TRUE)) {
  library(pins)
}
```

## Overview

This notebook is about board maintenance:

- inspecting local model pins
- publishing local model pins to another board
- checking versions and metadata
- deleting or replacing pins when needed

The stable model bundle now centers on:

- `checkpoint_best_total.pth`
- `manifest.json`
- `training_summary.json`

Older pins may also include `metadata.json`, `metrics.csv`, `hparams.yaml`,
`log.txt`, or `results.json`.

## Configure Boards

```{r}
# Local board used by petrographer training
board_local <- pins::board_folder(here(".petrographer/models"), versioned = TRUE)
board_local
```

```{r}
# Example shared folder board
# board_shared <- pins::board_folder("~/Dropbox/petrographer-models", versioned = TRUE)
```

```{r}
# Example S3 board
# board_s3 <- pins::board_s3(bucket = Sys.getenv("PETRO_S3_BUCKET"), prefix = "petrographer", versioned = TRUE)
```

## Inspect Local Models

```{r}
available <- if (requireNamespace("pins", quietly = TRUE)) pins::pin_list(board_local) else character()
available
```

```{r}
if (length(available)) {
  model_info(available[[1]], board = board_local)
}
```

```{r}
if (length(available)) {
  info <- pins::pin_meta(board_local, available[[1]])
  str(info$user, max.level = 2)
}
```

## Locate the Files Behind a Pin

Use `pin_download()` when you need the actual model directory for publishing or
inspection.

```{r}
if (length(available)) {
  files <- pins::pin_download(board_local, available[[1]])
  basename(files)
}
```

```{r}
if (exists("files")) {
  model_dir <- fs::path_dir(files[[1]])
  model_dir
}
```

## Publish a Local Pin to Another Board

`train_model()` already pins successful runs to the local board. Maintainers
only need `pin_model()` when copying a finished local pin to another board.

```{r}
if (FALSE && exists("model_dir")) {
  pin_model(
    model_dir = model_dir,
    model_id = available[[1]],
    board = board_shared,
    metadata = list(
      notes = "Published from model_zoo_pins.qmd",
      owner = Sys.info()[["user"]]
    )
  )
}
```

If the destination board is consumed through `board_url()`, refresh the board
manifest after publishing:

```{r}
if (FALSE) {
  pins::write_board_manifest(board_shared)
}
```

## Versions and Search

```{r}
if (length(available)) {
  pins::pin_versions(board_local, available[[1]])
}
```

```{r}
if (requireNamespace("pins", quietly = TRUE)) {
  res <- pins::pin_search(board_local, "shell")
  if (!is.null(res) && "name" %in% names(res)) unique(res$name) else character()
}
```

## Load a Model for a Quick Sanity Check

```{r}
if (length(available)) {
  mdl <- from_pretrained(available[[1]], board = "local", confidence = 0.5, device = "cpu")
  mdl
}
```

```{r}
# Example prediction (adjust the path to an existing image)
# if (exists("mdl") && file.exists("data/raw/sample.jpg")) {
#   res <- predict_image("data/raw/sample.jpg", mdl, use_slicing = TRUE)
#   head(res)
# }
```

## Housekeeping

Delete a pin only when you are sure you no longer need it on that board.

```{r}
if (FALSE && length(available) > 0) {
  pins::pin_delete(board_local, available[[1]])
}
```

## What to Reuse Later

This notebook is mainly worth keeping for:

- local/shared board configuration snippets
- publishing flow with `pin_model()`
- `pin_download()` plus `model_dir` extraction
- quick inspection of board metadata and versions
